---
title: "Combination designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combination designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignettes.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Here we show how `crmPack` can be used for drug combination dose escalation designs. We start with the simple case of a single combination arm, and then move on to the more complex case of parallel monotherapy and combination arms.

## Single combination arm

The package implements the 5-parameter BLRM approach by @Neuenschwander2015 (note that a presentation slide deck is available [here](https://de.slideshare.net/slideshow/eugm-2014-roychaudhuri-phase-1-combination/50032570)).

### Model specification

- Let $\textrm{odds}(p)=p/(1-p)$ be the odds transformation of the probability $p$, such that 
  $\textrm{logit}(p) = \log(\textrm{odds}(p))$. 
- Let $x_i$ be the dose of drug $i=1,2$, and $p(x_1, x_2)$ be the probability of DLT with doses $x_1$ and $x_2$.
- The reference doses for the two compounds are again denoted by $x_1^{*}$ and $x_2^{*}$, and the single-agent DLT probabilities at these reference doses are denoted by $p(x_1)$ and $p(x_2)$, respectively.
- Then the model assumes a linear interaction function:

  $$
  \textrm{odds}(p(x_1, x_2)) = \textrm{odds}(p_0(x_1, x_2)) \cdot \exp(\eta x_1/x_1^{*} x_2/x_2^{*}),
  $$

  - where $\eta$ is the interaction coefficient (positive values correspond to synergistic toxicity, zero corresponds to additive effect without interaction, and negative values correspond to antagonistic toxicity). 
  - This model is also called the "dual combo model" by Novartis
  - Under no interaction with $\eta=0$, this reduces the probability $\textrm{odds}(p(x_1, x_2))$ to
  \begin{align*}
  p_0(x_1, x_2) &= p(x_1) + p(x_2) - p(x_1)p(x_2)\\
   &= 1 - (1 - p(x_1))(1 - p(x_2)).
  \end{align*}
  - As a prior for $\eta$ we use either a normal distribution (if negative values, i.e. antagonistic effects, are plausible) or a log-normal distribution (if only positive values, i.e. synergistic effects, are plausible).

- Now for the single-agent DLT probabilities $p(x_1)$ and $p(x_2)$ we assume the standard logistic log-normal models:
  
  $$
  \textrm{logit}[p(x_i)] = \alpha_i + \beta_i \log(x_i/x_i^{*}),
  $$

  with prior e.g. $(\alpha_i, \log(\beta_i))^\top \sim \textrm{Normal}(\mu_i, \Sigma_i)$ for $i=1,2. 

### Model implementation

This model can be used in `crmPack` as follows with the `LogisticLogNormalCombo` model class:

```{r}
devtools::load_all()

combo_model <- LogisticLogNormalCombo(
  single_models = list(
    drug1 = LogisticLogNormal(
      mean = c(-0.85, 1),
      cov = matrix(c(2, -0.5, -0.5, 2), nrow = 2),
      ref_dose = 10
    ),
    drug2 = LogisticLogNormal(
      mean = c(-0.7, 0.8),
      cov = matrix(c(2, -0.3, -0.3, 2), nrow = 2),
      ref_dose = 20
    )
  ),
  gamma = 0,
  tau = 1
)

combo_model
```

### Design implementation

The corresponding design class is `DesignCombo`. Here is an example:

```{r}
my_next_best <- NextBestNCRM(
  target = c(0.2, 0.35),
  overdose = c(0.35, 1),
  max_overdose_prob = 0.25
)

my_increments <- IncrementsMin(
  increments_list = list(
    IncrementsComboOneDrugOnly(),
    IncrementsComboCartesian(
      drug1 = IncrementsRelative(intervals = c(0), increments = c(2)),
      drug2 = IncrementsRelative(intervals = c(0), increments = c(2))
    )
  )
)

my_stopping <- StoppingMinPatients(nPatients = 20)

empty_data <- DataCombo(
  doseGrid = list(
    drug1 = 10:30,
    drug2 = 20:60
  )
)

design_combo <- DesignCombo(
  model = combo_model,
  nextBest = my_next_best,
  stopping = my_stopping,
  increments = my_increments,
  cohort_size = CohortSizeConst(3L),
  data = empty_data,
  startingDose = c(10, 20)
)

design_combo
```

We can then simulate from this design as follows:

```{r}
true_tox_combo <- function(dose) {
  plogis(-4 + 0.08 * dose[1] + 0.06 * dose[2] + 0.001 * dose[1] * dose[2])
}

options <- McmcOptions(
  burnin = 100,
  step = 2,
  samples = 100,
  rng_kind = "Mersenne-Twister",
  rng_seed = 1
)

sims_combo <- simulate(
  design_combo,
  truth = true_tox_combo,
  nsim = 20,
  seed = 819,
  mcmcOptions = options,
  parallel = FALSE
)

plot(sims_combo)
summary(sims_combo, truth = true_tox_combo)
```

## References