---
title: "Combination designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Combination designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignettes.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

Here we show how `crmPack` can be used for drug combination dose escalation designs. We start with the simple case of a single combination arm, and then move on to the more complex case of parallel monotherapy and combination arms.

## Single combination arm

The package implements the 5-parameter BLRM approach by @Neuenschwander2015 (note that a presentation slide deck is available [here](https://de.slideshare.net/slideshow/eugm-2014-roychaudhuri-phase-1-combination/50032570)).

### Model specification

- Let $\textrm{odds}(p)=p/(1-p)$ be the odds transformation of the probability $p$, such that 
  $\textrm{logit}(p) = \log(\textrm{odds}(p))$. 
- Let $x_i$ be the dose of drug $i=1,2$, and $p(x_1, x_2)$ be the probability of DLT with doses $x_1$ and $x_2$.
- The reference doses for the two compounds are again denoted by $x_1^{*}$ and $x_2^{*}$, and the single-agent DLT probabilities at these reference doses are denoted by $p(x_1)$ and $p(x_2)$, respectively.
- Then the model assumes a linear interaction function:

  $$
  \textrm{odds}(p(x_1, x_2)) = \textrm{odds}(p_0(x_1, x_2)) \cdot \exp(\eta x_1/x_1^{*} x_2/x_2^{*}),
  $$

  - where $\eta$ is the interaction coefficient (positive values correspond to synergistic toxicity, zero corresponds to additive effect without interaction, and negative values correspond to antagonistic toxicity). 
  - This model is also called the "dual combo model" by Novartis
  - Under no interaction with $\eta=0$, this reduces the probability $\textrm{odds}(p(x_1, x_2))$ to
  \begin{align*}
  p_0(x_1, x_2) &= p(x_1) + p(x_2) - p(x_1)p(x_2)\\
   &= 1 - (1 - p(x_1))(1 - p(x_2)).
  \end{align*}
  - As a prior for $\eta$ we use either a normal distribution (if negative values, i.e. antagonistic effects, are plausible) or a log-normal distribution (if only positive values, i.e. synergistic effects, are plausible).

- Now for the single-agent DLT probabilities $p(x_1)$ and $p(x_2)$ we assume the standard logistic log-normal models:
  
  $$
  \textrm{logit}[p(x_i)] = \alpha_i + \beta_i \log(x_i/x_i^{*}),
  $$

  with prior e.g. $(\alpha_i, \log(\beta_i))^\top \sim \textrm{Normal}(\mu_i, \Sigma_i)$ for $i=1,2. 

### Model implementation

This model can be used in `crmPack` as follows with the `TwoDrugsCombo` model class:

```{r}
devtools::load_all()

combo_model <- TwoDrugsCombo(
  single_models = list(
    drug1 = LogisticLogNormal(
      mean = c(-0.85, 1),
      cov = matrix(c(2, -0.5, -0.5, 2), nrow = 2),
      ref_dose = 10
    ),
    drug2 = LogisticLogNormal(
      mean = c(-0.7, 0.8),
      cov = matrix(c(2, -0.3, -0.3, 2), nrow = 2),
      ref_dose = 20
    )
  ),
  gamma = 0,
  tau = 1
)

combo_model
```

### Design implementation

The corresponding design class is `DesignCombo`. Here is an example:

```{r}
my_next_best <- NextBestNCRM(
  target = c(0.2, 0.35),
  overdose = c(0.35, 1),
  max_overdose_prob = 0.25
)

my_increments <- IncrementsMin(
  increments_list = list(
    IncrementsComboOneDrugOnly(),
    IncrementsComboCartesian(
      drug1 = IncrementsRelative(intervals = c(0), increments = c(2)),
      drug2 = IncrementsRelative(intervals = c(0), increments = c(2))
    )
  )
)

my_stopping <- StoppingMinPatients(nPatients = 20)

empty_data <- DataCombo(
  doseGrid = list(
    drug1 = seq(from = 10, to = 50, by = 5),
    drug2 = seq(from = 10, to = 50, by = 5)
  )
)

design_combo <- DesignCombo(
  model = combo_model,
  nextBest = my_next_best,
  stopping = my_stopping,
  increments = my_increments,
  cohort_size = CohortSizeConst(3L),
  data = empty_data,
  startingDose = c(drug1 = 10, drug2 = 20)
)

design_combo
```

We can then simulate from this design as follows:

```{r}
true_tox_combo <- function(dose) {
  plogis(-4 + 0.08 * dose[1] + 0.06 * dose[2] + 0.001 * dose[1] * dose[2])
}

mcmc_options <- McmcOptions(
  burnin = 100,
  step = 2,
  samples = 100,
  rng_kind = "Mersenne-Twister",
  rng_seed = 1
)

sims_combo <- simulate(
  design_combo,
  truth = true_tox_combo,
  nsim = 20,
  seed = 819,
  mcmcOptions = mcmc_options,
  parallel = FALSE
)

plot(sims_combo)
summary(sims_combo, truth = true_tox_combo)
```

We can also look at the trajectory more closely:

```{r}
plot(sims_combo, type = "trajectory2D")
```

## Parallel monotherapy and combination arms

With `crmPack` we can also implement more complex designs with parallel monotherapy and combination arms. This is done with the "hierarchical" classes, as we will show now.

### Model specification

We use logistic log normal models for the monotherapy arm parameters, and the same BLRM model for the combination arm parameters as described above. The parameters are then linked via common hyperprior distributions:

- We use a meta-analytic combined (MAC) prior which separates according to different trial sources:
  - mono drug 1 trial ($\theta_1$), which will only inform the parameters $\alpha_1$ and $\beta_1$ for drug 1
  - current combination of drugs 1 and 2 trial ($\theta_2$), which will inform all 5 parameters
  - mono drug 2 trial ($\theta_3$), which will only inform the parameters $\alpha_2$ and $\beta_2$ for drug 2
- On top of these sits the hierarchical prior $\theta_j \sim \textrm{Normal}(\mu, \Sigma)$, $j=1,2,3$.
- Hyperpriors on $\mu = (\mu_{\alpha_1}, \mu_{\beta_1}, \mu_{\alpha_2}, \mu_{\beta_2}, \mu_{\eta})^\top$ e.g.:
  - $\mu_{\alpha_i} \sim \textrm{Normal}(\textrm{logit}(0.25), 2.5^2)$, $i=1,2$
  - $\mu_{\beta_1} \sim \textrm{Normal}(0, 0.7^2)$
  - $\mu_{\beta_2}, \mu_{\eta} \sim \textrm{Normal}(0, 1)$
- Structured assumptions for $\Sigma$
  - Assume that only $\alpha_{1}$ and $\beta_1$, as well as $\alpha_{2}$ and $\beta_2$ are correlated, but not the other parameters, then we have a corresponding block diagonal structure for $\Sigma$ with 7 parameters
  (standard deviations $\tau_{\alpha_1}, \tau_{\beta_1}, \tau_{\alpha_2}, \tau_{\beta_2}, \tau_{\eta}$ for each of the 5 parameters plus the two correlations $\rho_1$ and $\rho_2$ between $\alpha_i$ and $\beta_i$ for $i=1,2$).
  - We don't fix the $\Sigma$ entries but instead again assign hyperpriors, e.g. with $\kappa = \log(2)/1.96$
    - $\tau_{\alpha_1} \sim \textrm{LogNormal}(\log(0.5), \kappa^2)$
    - $\tau_{\alpha_2} \sim \textrm{LogNormal}(\log(0.75), \kappa^2)$
    - $\tau_{\beta_1}, \tau_{\beta_2} \sim \textrm{LogNormal}(\log(0.25), \kappa^2)$
    - $\tau_{\eta} \sim \textrm{LogNormal}(\log(0.125), \kappa^2)$
    - $\rho_1, \rho_2 \sim \textrm{Uniform}(-1, 1)$

### Model implementation

This model can be implemented in `crmPack` with the `HierarchicalModel` class as follows:

```{r}
hierarchical_model <- HierarchicalModel(
    mono_drug1 = combo_model@single_models$drug1,
    mono_drug2 = combo_model@single_models$drug2,
    combo = combo_model,
    exchangeable_parameters = list(
        drug1_intercept = list(
            mono_drug1 = "alpha0",
            combo = "alpha0[1]"
        ),
        drug1_slope = list(
            mono_drug1 = "alpha1",
            combo = "alpha1[1]"
        ),
        drug2_intercept = list(
            mono_drug2 = "alpha0",
            combo = "alpha0[2]"
        ),
        drug2_slope = list(
            mono_drug2 = "alpha1",
            combo = "alpha1[2]"
        )
    )
)

hierarchical_model
```

We can look at the JAGS code, which has been automatically compiled for this model.

The likelihood is as follows:

```{r}
hierarchical_model@datamodel
```

The prior is as follows:

```{r}
hierarchical_model@priormodel
```

### Design implementation

The corresponding design can be implemented with the `HierarchicalDesign` class as follows. 
First, we define the arm-specific designs, and wrap them with `DesignArm` to specify their names and whether they are active or not (which would be useful for historical data from previous trials).

```{r}
design_mono <- Design(
  model = LogisticLogNormal(
      mean = c(-0.85, 1),
      cov = matrix(c(2, -0.5, -0.5, 2), nrow = 2),
      ref_dose = 10
    ),
  nextBest = my_next_best,
  stopping = StoppingMinPatients(nPatients = 20) | StoppingMissingDose(),
  increments = IncrementsRelative(intervals = c(0), increments = c(2)),
  cohort_size = CohortSizeConst(3L),
  data = Data(
    doseGrid = seq(from = 10, to = 50, by = 5)
  ),
  startingDose = 10
)

design_arm_mono <- DesignArm(
  name = "mono",
  design = design_mono
)
design_arm_combo <- DesignArm(
  name = "combo",
  design = design_combo
)
```

Note that for simplicity here we just have a single mono arm.
Then we provide these to the constructor, together with the specification of the exchangeable parameters (in the same way as above for the model):

```{r}
hierarchical_design <- HierarchicalDesign(
  design_arm_mono,
  design_arm_combo,
  exchangeable_parameters = list(
    drug1_intercept = list(
      mono = "alpha0",
      combo = "alpha0[1]"
    ),
    drug1_slope = list(
      mono = "alpha1",
      combo = "alpha1[1]"
    )
  )
)

hierarchical_design
```

Here we basically also show that the mono arm is about `drug1` and not `drug2`, which is relevant for the information sharing via the exchangeable parameters.

Then we can simulate from this design as follows:

```{r}
truth_mono <- function(dose) plogis(-4 + 0.08 * dose)
truth_combo <- function(dose) plogis(-4 + 0.08 * dose[1] + 0.06 * dose[2] + 0.001 * dose[1] * dose[2])

hierarchical_sims <- simulate(
  hierarchical_design,
  truth = list(
    mono = truth_mono,
    combo = truth_combo
  ),
  truthResponse = list(
    # Just for demo purposes we choose the same functions for efficacy.
    mono = truth_mono,
    combo = truth_combo
  ),
  nsim = 20,
  seed = 819,
  mcmcOptions = mcmc_options,
  parallel = FALSE
)

hierarchical_sims
summary(hierarchical_sims, truth = list(
  mono = truth_mono,
  combo = truth_combo
))
```

It is also possible to extract the arm specific simulations to produce separate plots etc.:

```{r}
arm_mono_sims <- get_arm_simulations(hierarchical_sims, "mono")
plot(arm_mono_sims)
```

## References
