---
title: "Comparison with the decider package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with the decider package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignettes.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we want to compare the combination design implemented in `crmPack` and explained in this [vignette](combo_designs.Rmd) with the implementation in the [`decider` package](https://boehringer-ingelheim.github.io/decider/) [@Schroeter2023]. 

## Example

We are going to use the example as described in the `decider` vignette [here](https://boehringer-ingelheim.github.io/decider/articles/intro_jointBLRM.html#setting-up-and-evaluating-priors-using-scenario_jointblrm):

- Three arms:
  - Arm A: monotherapy of compound 1
  - Arm B: combination of compound 1 and compound 2
  - Arm C: historical data from compound 2
- Arm B can start when certain doses of Arm A have been cleared
- Logistic log-normal models for compound 1 and compound 2
- Target interval is 16-33% DLT rate
- Prior specification
  - uniform prior for the correlation between intercept and log-slope in the logistic log-normal model
  - for the hyper-means (i.e. mean of parameters across trials)
  - between-trial heterogeneity (i.e. standard deviation of parameters across trials)


## Using `decider`

```{r}
library(decider)
```

This is the data from the historical Arm C:

```{r}
historical_data <- list(
  dose1 = c( 0,    0,    0,    0,    0),
  dose2 = c( 2,    4,    8,    12,   16),
  n.pat = c( 3,    3,    3,    9,    12),
  n.dlt = c( 0,    0,    0,    1,    2),
  trial = c("H1", "H1", "H1", "H1", "H1")
)
```

The monotherapy dose grid for Arm A is:

```{r}
d1 <- c(0.1, 0.2, 0.4, 0.8, 1.6, 2.4, 3.6, 5, 6)
```

The dose grid for compound 2 in Arm B is more sparse:

```{r}
d2 <- c(8, 12)
```

The overall dose grid for combination Arm B is therefore:

```{r}
doses_of_interest <- rbind( 
    c(d1,                  rep(d1, times=length(d2))),
    c(rep(0, length(d1)),  rep(d2, each=length(d1))) 
)
```

The reference doses to be used in the models are:

```{r}
dose_ref1 <- 6
dose_ref2 <- 12
```

We further need to specify the arms and types of the arms as follows:

```{r}
trials_of_interest = c("A",     "B")
types_of_interest =  c("mono1", "combi")
```

The prior for the hypermeans is specified like this:

```{r}
#                Parameter   Mean         SD
prior_mu <- list(mu_a1  =  c(logit(0.33), 2),
                 mu_b1  =  c(0,           1), # standard normal
                 mu_a2  =  c(logit(0.33), 2),
                 mu_b2  =  c(0,           1), # standard normal
                 mu_eta =  c(0,           1.121))
```

The prior mean for $\mu_{\alpha_{1}}$ is set to $\text{logit}(0.33)$, which implies that we assume the reference dose has a prior median DLT rate of 33%.

Note that we use a normal prior here on the interaction parameter $\eta$, thus allowing both positive and negative interactions. The standard deviation is set such that $\exp(1.96 \cdot 1.121) \approx 9$, thus allowing for a 95% prior interval of $[1/9, 9]$ for the odds changes for a DLT at the reference dose. So $1.121 = \log(9)/z_{0.975}$.

The prior for the between-trial heterogeneity parameters is specified like this:

```{r}
#                 Parameter    Mean        SD
prior_tau <- list(tau_a1  =  c(log(0.25),  log(2)/1.96),
                  tau_b1  =  c(log(0.125), log(2)/1.96),
                  tau_a2  =  c(log(0.25),  log(2)/1.96),
                  tau_b2  =  c(log(0.125), log(2)/1.96),
                  tau_eta =  c(log(0.125), log(2)/1.96))
```

These are all the log normal prior parameters for the corresponding $\tau$ parameters. These are all "moderate" degrees of heterogeneity, according to @Neuenschwander2015.

Then we look at the following scenario, where two cohorts of patients are available from Arm A:

```{r}
scenario1 <- list(
  dose1 = c( 0.1,  0.2),
  dose2 = c( 0,    0),
  n.pat = c( 3,    3),
  n.dlt = c( 0,    1),
  trial = c("A",  "A")
)
```

We note that the `trial` specification here needs to match the name used in `trials_of_interest` above.

Now we can call the scenario function:

```{r}
result1 <- scenario_jointBLRM(
  data = scenario1,
  historical.data = historical_data,
  doses.of.interest = doses_of_interest,
  dose.ref1 = dose_ref1,
  dose.ref2 = dose_ref2,
  trials.of.interest = trials_of_interest,
  types.of.interest = types_of_interest,
  prior.mu = prior_mu,
  prior.tau = prior_tau
)
```

We can look at the results:

```{r}
result1
```

For each trial of interest, the posterior toxicities previously designated to be of interest are shown.

Under the hood, the implementation works as follows:

- [`post_tox_jointBLRM()`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/sampling_jointBLRM.R#L232) is called to sample from the posterior, which in turn uses
- [`sampling_jointBLRM()`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/sampling_jointBLRM.R#L17) which then calls `rstan::sampling()` on
- [`stanmodels$jointBLRM`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/stanmodels.R#L11) which is the constant Stan model sourced from
- [`jointBLRM.stan`](https://github.com/Boehringer-Ingelheim/decider/blob/main/inst/stan/jointBLRM.stan)

So we can compare this with the implementation in `crmPack` which is based on JAGS.

## Using `crmPack`

Now we are going to define the same design and scenario in `crmPack`. 

We start with the monotherapy model for compound 1:

```{r}
devtools::load_all() # library(crmPack)

mono_model1 <- LogisticLogNormal(
  mean = c(logit(0.33), 0),
  cov = diag(c(2, 1)^2),
  ref_dose = dose_ref1
)
```

And for compound 2 the same:

```{r}
mono_model2 <- LogisticLogNormal(
    mean = c(logit(0.33), 0),
    cov = diag(c(2, 1)^2),
    ref_dose = dose_ref2
)
```

Then we define the combination model:

```{r}
combo_model <- TwoDrugsCombo(
    list(
        compound1 = mono_model1,
        compound2 = mono_model2
    ),
    gamma = 0, # prior mean for the interaction parameter
    tau = 1 / (1.121^2) # prior precision for the interaction parameter
)
```

We define the historical data which is already available:

```{r}
hist_data_comp2 <- Data(
    x = rep(historical_data$dose2, historical_data$n.pat),
    y = c(
        rep(0, sum(historical_data$n.pat) - sum(historical_data$n.dlt)), 
        rep(1, sum(historical_data$n.dlt))
    ),
    doseGrid = historical_data$dose2
)
```

We are going to use simple rules here (they are not relevant for the current scenario comparison):

```{r}
my_stopping <- StoppingMinPatients(nPatients = 50)
my_increments <- IncrementsRelative(0, 2)
my_nextBest <- NextBestNCRM(
    target = c(0.16, 0.33), 
    overdose = c(0.33, 1), 
    max_overdose_prob = 0.25
)
my_cohort_size <- CohortSizeConst(size = 3)
my_increments_combo <- IncrementsComboOneDrugOnly()
```

Then we define the design arms accordingly:

```{r}
design_armA <- DesignArm(
    "A",
    active = TRUE,
    design = Design(
        data = Data(doseGrid = d1),
        startingDose = d1[1],
        model = mono_model1,
        stopping = my_stopping,
        increments = my_increments,
        nextBest = my_nextBest,
        cohort_size = my_cohort_size
    )
)

design_armB <- DesignArm(
    "B",
    active = TRUE,
    design = DesignCombo(
        data = DataCombo(doseGrid = list(compound1 = d1, compound2 = c(0, d2))),
        startingDose = c(compound1 = d1[1], compound2 = 0),
        model = combo_model,
        stopping = my_stopping,
        increments = my_increments_combo,
        nextBest = my_nextBest,
        cohort_size = my_cohort_size
    ),
    open_when = ArmMinDoseCondition("A", min_dose = d1[2])
)

design_armC <- DesignArm(
    "C",
    active = FALSE,
    data = hist_data_comp2,
    model = mono_model2
)
```

Now we can define the hierarchical design:

```{r}
design_hierarchical <- HierarchicalDesign(
    design_armA,
    design_armB,
    design_armC,
    exchangeable_parameters = list(
        comp1_intercept = list(
            A = "alpha0",
            B = "alpha0[1]"
        ),
        comp1_slope = list(
            A = "alpha1",
            B = "alpha1[1]"
        ),
        comp2_intercept = list(
            B = "alpha0[2]",
            C = "alpha0"
        ),
        comp2_slope = list(
            B = "alpha1[2]",
            C = "alpha1"    
        )
    ),
    pool_correlations = list(
        comp1 = c("comp1_intercept", "comp1_slope"),
        comp2 = c("comp2_intercept", "comp2_slope")
    ),
    pool_priors = list(
        comp1_intercept = list(
            mu = prior_mu$mu_a1,
            tau = prior_tau$tau_a1
        ),
        comp1_slope = list(
            mu = prior_mu$mu_b1,
            tau = prior_tau$tau_b1
        ),
        comp2_intercept = list(
            mu = prior_mu$mu_a2,
            tau = prior_tau$tau_a2
        ),
        comp2_slope = list(
            mu = prior_mu$mu_b2,
            tau = prior_tau$tau_b2
        )
    )
)
```

Note that each entry in `pool_correlations` can correlate exactly two scalar
exchangeable parameter pools. In this example, `comp1` correlates the compound 1
intercept pool with the compound 1 slope pool, and `comp2` does the same for
compound 2. Correlated blocks with three or more parameters are not currently
supported.

Then we define the scenario:

```{r}
scenario_hierarchical <- HierarchicalData(
    A = Data(
        x = c(0.1, 0.1, 0.1, 0.2, 0.2, 0.2),
        y = c(0, 0, 0, 0, 0, 1),
        doseGrid = design_armA@design@data@doseGrid
    ),
    B = design_armB@design@data,
    C = design_armC@design@data
)
```

And then we can use the `scenario()` function:

```{r}
result1_crmPack <- scenario(
    design_hierarchical,
    data = scenario_hierarchical,
    mcmcOptions = McmcOptions()
)
```

We can look at the fit results:

```{r}
result1_crmPack$fit
```

We can also check the probabilities to be in target and overdosing intervals:

```{r}
result1_crmPack$next_best$A$probs
result1_crmPack$next_best$B$probs
```

## Comparison of fit

Based on this we can first compare the fit results.

Let's look at the results for Arm A:

```{r}
fit_trialA_decider <- result1$`trial-A` |> as.data.frame()
fit_trialA_crmPack <- result1_crmPack$fit$A
probs_trialA_crmPack <- result1_crmPack$next_best$A$probs |> as.data.frame()
diff_trialA <- data.frame(
    dose = fit_trialA_crmPack$dose,
    center = fit_trialA_decider$mean - fit_trialA_crmPack$middle,
    lower = fit_trialA_decider$`q.2.5%` - fit_trialA_crmPack$lower,
    upper = fit_trialA_decider$`q.97.5%` - fit_trialA_crmPack$upper,
    target = fit_trialA_decider$`P([0.16,0.33))` - probs_trialA_crmPack$target,
    overdose = fit_trialA_decider$`P([0.33,1])` - probs_trialA_crmPack$overdose
)
diff_trialA
```

And then the results for Arm B:

```{r}
fit_trialB_decider <- result1$`trial-B` |> as.data.frame()
fit_trialB_crmPack <- result1_crmPack$fit$B |> dplyr::filter(compound2 > 0)
probs_trialB_crmPack <- result1_crmPack$next_best$B$probs |> as.data.frame() |> dplyr::filter(compound2 > 0) 
diff_trialB <- data.frame(
    dose1 = fit_trialB_crmPack$compound1,
    dose2 = fit_trialB_crmPack$compound2,
    center = fit_trialB_decider$mean - fit_trialB_crmPack$middle,
    lower = fit_trialB_decider$`q.2.5%` - fit_trialB_crmPack$lower,
    upper = fit_trialB_decider$`q.97.5%` - fit_trialB_crmPack$upper,
    target = fit_trialB_decider$`P([0.16,0.33))` - probs_trialB_crmPack$target,
    overdose = fit_trialB_decider$`P([0.33,1])` - probs_trialB_crmPack$overdose
)
diff_trialB
```

So these differences look relatively small, and there does not seem to be any 
systematic bias in the differences.

## Comparison of model code

Let's compare the model code used in `decider` and `crmPack`, in order to make sure that they really match and implement the same priors and models:

### `decider`

Here we have the following Stan model:

```{r}
#| results: "asis"
#| echo: false

cat("````\n")
cat(decider:::stanmodels$jointBLRM@model_code)
cat("\n````")
```

### `crmPack`

Here we have the following JAGS model:

```{r}
#| results: "asis"
#| echo: false

cat("````", deparse(body(design_hierarchical@model@datamodel)), "````", sep = "\n")

cat("````", deparse(body(design_hierarchical@model@priormodel)), "````", sep = "\n")
```

### Conclusion

There are still some minor differences:

- In `decider`, the $\eta$ parameter is part of the 5-parameter hierarchical vector and has its own hypermean and heterogeneity. In `crmPack`, the $\eta$ parameter is not part of the hierarchical vector. When there is only combo arm then this does not matter.
- JAGS uses Bernoulli observations, Stan uses binomial cohort counts, but the likelihoods are equivalent.

Apart from these, the probabilistic models are equivalent. We could also see that in the fit results which are very close to each other.

## References
