---
title: "Comparison with the decider package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with the decider package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignettes.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we want to compare the combination design implemented in `crmPack` and explained in this [vignette](combo_designs.Rmd) with the implementation in the [`decider` package](https://boehringer-ingelheim.github.io/decider/) [@Schroeter2023]. 

## Example

We are going to use the example as described in the `decider` vignette [here](https://boehringer-ingelheim.github.io/decider/articles/intro_jointBLRM.html#setting-up-and-evaluating-priors-using-scenario_jointblrm):

- Three arms:
  - Arm A: monotherapy of compound 1
  - Arm B: combination of compound 1 and compound 2
  - Arm C: historical data from compound 2
- Arm B can start when certain doses of Arm A have been cleared
- Logistic log-normal models for compound 1 and compound 2
- Target interval is 16-33% DLT rate
- Prior specification
  - uniform prior for the correlation between intercept and log-slope in the logistic log-normal model
  - for the hyper-means (i.e. mean of parameters across trials)
  - between-trial heterogeneity (i.e. standard deviation of parameters across trials)


## Using `decider`

```{r}
library(decider)
```

This is the data from the historical Arm C:

```{r}
historical_data <- list(
  dose1 = c( 0,    0,    0,    0,    0),
  dose2 = c( 2,    4,    8,    12,   16),
  n.pat = c( 3,    3,    3,    9,    12),
  n.dlt = c( 0,    0,    0,    1,    2),
  trial = c("H1", "H1", "H1", "H1", "H1")
)
```

The monotherapy dose grid for Arm A is:

```{r}
d1 <- c(0.1, 0.2, 0.4, 0.8, 1.6, 2.4, 3.6, 5, 6)
```

The dose grid for compound 2 in Arm B is more sparse:

```{r}
d2 <- c(8, 12)
```

The overall dose grid for combination Arm B is therefore:

```{r}
doses_of_interest <- rbind( 
    c(d1,                  rep(d1, times=length(d2))),
    c(rep(0, length(d1)),  rep(d2, each=length(d1))) 
)
```

The reference doses to be used in the models are:

```{r}
dose_ref1 <- 6
dose_ref2 <- 12
```

We further need to specify the arms and types of the arms as follows:

```{r}
trials_of_interest = c("A",     "B")
types_of_interest =  c("mono1", "combi")
```

The prior for the hypermeans is specified like this:

```{r}
#                Parameter   Mean         SD
prior_mu <- list(mu_a1  =  c(logit(0.33), 2),
                 mu_b1  =  c(0,           1), # standard normal
                 mu_a2  =  c(logit(0.33), 2),
                 mu_b2  =  c(0,           1), # standard normal
                 mu_eta =  c(0,           1.121))
```

The prior mean for $\mu_{\alpha_{1}}$ is set to $\text{logit}(0.33)$, which implies that we assume the reference dose has a prior median DLT rate of 33%.

Note that we use a normal prior here on the interaction parameter $\eta$, thus allowing both positive and negative interactions. The standard deviation is set such that $\exp(1.96 \cdot 1.121) \approx 9$, thus allowing for a 95% prior interval of $[1/9, 9]$ for the odds changes for a DLT at the reference dose. So $1.121 = \log(9)/z_{0.975}$.

The prior for the between-trial heterogeneity parameters is specified like this:

```{r}
#                 Parameter    Mean        SD
prior_tau <- list(tau_a1  =  c(log(0.25),  log(2)/1.96),
                  tau_b1  =  c(log(0.125), log(2)/1.96),
                  tau_a2  =  c(log(0.25),  log(2)/1.96),
                  tau_b2  =  c(log(0.125), log(2)/1.96),
                  tau_eta =  c(log(0.125), log(2)/1.96))
```

These are all the log normal prior parameters for the corresponding $\tau$ parameters. These are all "moderate" degrees of heterogeneity, according to @Neuenschwander2015.

Then we look at the following scenario, where two cohorts of patients are available from Arm A:

```{r}
scenario1 <- list(
  dose1 = c( 0.1,  0.2),
  dose2 = c( 0,    0),
  n.pat = c( 3,    3),
  n.dlt = c( 0,    1),
  trial = c("A",  "A")
)
```

We note that the `trial` specification here needs to match the name used in `trials_of_interest` above.

Now we can call the scenario function:

```{r}
result1 <- scenario_jointBLRM(
  data = scenario1,
  historical.data = historical_data,
  doses.of.interest = doses_of_interest,
  dose.ref1 = dose_ref1,
  dose.ref2 = dose_ref2,
  trials.of.interest = trials_of_interest,
  types.of.interest = types_of_interest,
  prior.mu = prior_mu,
  prior.tau = prior_tau
)
```

We can look at the results:

```{r}
result1
```

For each trial of interest, the posterior toxicities previously designated to be of interest are shown.

Under the hood, the implementation works as follows:

- [`post_tox_jointBLRM()`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/sampling_jointBLRM.R#L232) is called to sample from the posterior, which in turn uses
- [`sampling_jointBLRM()`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/sampling_jointBLRM.R#L17) which then calls `rstan::sampling()` on
- [`stanmodels$jointBLRM`](https://github.com/Boehringer-Ingelheim/decider/blob/main/R/stanmodels.R#L11) which is the constant Stan model sourced from
- [`jointBLRM.stan`](https://github.com/Boehringer-Ingelheim/decider/blob/main/inst/stan/jointBLRM.stan)

So we can compare this with the implementation in `crmPack` which is based on JAGS.

## References