---
title: "Describing `crmPack` Objects"
output: rmarkdown::html_vignette
editor: visual
number-sections: true
embed-resources: TRUE
reference-location: margin
citation-location: margin
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{demo}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  error: true
  warning: true
  eval: false
---

```{r}
#| label: setup
#| echo: FALSE

suppressPackageStartupMessages({
  library(tibble)
  library(dplyr)
  library(knitr)
  library(kableExtra)
  library(crmPack)
})
```

## Introduction

Objects created by `crmPack` are almost always S4 objects. Like all S4 objects, by default they do
not render in a particularly user-friendly way.

```{r}
#| label: not-friendly-1
#| echo: TRUE
#| eval: FALSE
cs <- CohortSizeDLT(intervals = 0:2, cohort_size = c(1, 3, 5))
cs
```

```{r}
#| label: not-friendly-2
#| echo: FALSE
#| eval: TRUE
cs <- CohortSizeDLT(intervals = 0:2, cohort_size = c(1, 3, 5))
normal_print(cs)
```

Fortunately, a little known feature of `knitr` can put this right at little or no cost to the end
user: in the simplest case, demonstrated below, all that needs to be done is to reference the object
in a markdown or Quarto chunk.

```{r}
#| label: not-friendly-3
#| echo: TRUE
#| eval: TRUE

cs
```

> The `knit_print`methods provided by `crmPack` are not intended to be fully customisable or
> comprehensive. We do believe, however, that they cover the vast majority of use-cases and are
> easily extended using the techniques described in @sec-override of this vignette.

> Formatting of these objects currently only works for HMTL output. If another format - such as PDF
> or Microsoft Word - is required, our suggested workaround is to create the HTML output and then
> print or save the document to the required format.
