---
title: "Migrating from crmPack version 1.0.3"
author: "Wojciech Wójciak"
date: "05 October 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from crmPack version 1.0.3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{crmPack}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(kableExtra)
library(checkmate)
```

The following Vignette describes the changes that were introduced to the `crmPack`
package as a result of the package's refactoring. One row in below tables
represents a single-type, consistent change.

# Class and Methods Name Changes

## Classes

```{r class_change_helpers, echo=FALSE}

# A simple helper function that generates change class record.
ch_class <- function(name = "", new_name = "", comment = "", add_constructor_info = TRUE) {
  assert_string(name)
  assert_string(new_name)
  assert_string(comment)
  assert_flag(add_constructor_info)

  was_removed <- ifelse(nchar(new_name) == 0, TRUE, FALSE)
  if (add_constructor_info && !was_removed) {
    constructor_info <- "\\+ constructor corresponding update"
    comment <- ifelse(
      nchar(comment) == 0,
      constructor_info,
      paste0(constructor_info, "; ", comment)
    )
  }
  list(name, new_name, "", "", was_removed, comment)
}

# A simple helper function that generates change slot record.
ch_slot <- function(class_name,
                    name = "",
                    new_name = "",
                    comment = "",
                    add_constructor_info = TRUE) {
  assert_string(class_name)
  assert_string(name)
  assert_string(new_name)
  assert_string(comment)
  assert_flag(add_constructor_info)

  was_removed <- ifelse(nchar(new_name) == 0, TRUE, FALSE)
  if (add_constructor_info) {
    constructor_info <- "\\+ constructor corresponding update"
    comment <- ifelse(
      nchar(comment) == 0,
      constructor_info,
      paste0(constructor_info, "; ", comment)
    )
  }
  list(class_name, "", name, new_name, was_removed, comment)
}
```


```{r class_changes, echo=FALSE}

class_changes <- tribble(
  ~class_name, ~class_new_name, ~slot_name, ~slot_new_name, ~was_removed, ~comment,
  !!!ch_class("", "positive_number", "to handle strictly positive valued slots
  (e.g. `ref_dose` in many model classes)", FALSE),
  !!!ch_class("", "ModelLogNormal", "new parent class for all the models with
  reference dose and bivariate (log) normal prior on the model parameters", FALSE),
  !!!ch_class("", "ModelParamsNormal", "represents parameters of bivariate normal distribution", FALSE),
  !!!ch_class("AllModels"),
  !!!ch_class("Model"),
  !!!ch_slot("GeneralModel", "", "datanames", "moved from `AllModels`", FALSE),
  !!!ch_slot("GeneralModel", "", "datanames_prior", add_constructor_info = FALSE),
  !!!ch_slot("ModelEff", "dose", "", "moved to model class method", FALSE),
  !!!ch_slot("ModelEff", "ExpEff", "", "moved to model class method and renamed to `efficacy`", FALSE),
  !!!ch_slot("Effloglog", "several slots", "several slots", "several slot changed, see man page"),
  !!!ch_slot("EffFlexi", "several slots", "several slots", "several slot changed, see man page"),
  !!!ch_slot("EffFlexi()", "smooth", "", "replaced by `rw1` flag", FALSE),
  !!!ch_slot("EffFlexi()", "", "rw1", "replaces `smooth` string", FALSE),
  !!!ch_slot("DALogisticLogNormal", "C_par", "c_par"),
  !!!ch_slot("DALogisticLogNormal", "conditionalPEM", "cond_pem"),
  !!!ch_slot("TITELogisticLogNormal", "weightMethod", "weight_method"),
  !!!ch_slot("OneParExpNormalPrior", "skeletonFun", "skel_fun", add_constructor_info = FALSE),
  !!!ch_slot("OneParExpNormalPrior", "", "skel_fun_inv", add_constructor_info = FALSE),
  !!!ch_slot("OneParExpNormalPrior", "skeletonProbs", "skel_probs"),
  !!!ch_slot("OneParExpNormalPrior()", "doseGrid", "dose_grid", add_constructor_info = FALSE),
  !!!ch_slot("DualEndpointBeta", "refDoseBeta", "ref_dose_beta"),
  !!!ch_slot("DualEndpointEmax", "refDoseEmax", "ref_dose_emax"),
  !!!ch_slot("DualEndpointRW", "useRW1", "rw1", add_constructor_info = FALSE),
  !!!ch_slot("DualEndpointRW()", "smooth", "", "replaced by `rw1` flag", FALSE),
  !!!ch_slot("DualEndpointRW()", "", "rw1", "replaces `smooth` string", FALSE),
  !!!ch_slot("McmcOptions", "", "rng_kind", "to be used by Random Number Generator in rJAGS"),
  !!!ch_slot("McmcOptions", "", "rng_seed", "to be used by Random Number Generator in rJAGS"),
  !!!ch_slot("NextBestTD", "targetDuringTrial", "prob_target_drt"),
  !!!ch_slot("NextBestTD", "targetEndOfTrial", "prob_target_eot"),
  !!!ch_slot("NextBestTDsamples", "targetDuringTrial", "prob_target_drt"),
  !!!ch_slot("NextBestTDsamples", "targetEndOfTrial", "prob_target_eot"),
  !!!ch_slot("NextBestNCRM", "maxOverdoseProb", "max_overdose_prob"),
  !!!ch_slot("NextBestDualEndpoint", "scale", "", "replaced by `target_relative` flag"),
  !!!ch_slot("NextBestDualEndpoint", "", "target_relative", "replaces `scale` string"),
  !!!ch_slot("NextBestDualEndpoint", "maxOverdoseProb", "max_overdose_prob"),
  !!!ch_slot("NextBestDualEndpoint", "targetThresh", "target_thresh"),
  !!!ch_slot("NextBestMaxGain", "DLEDuringTrialtarget", "prob_target_drt"),
  !!!ch_slot("NextBestMaxGain", "DLEEndOfTrialtarget", "prob_target_eot"),
  !!!ch_slot("NextBestMaxGainSamples", "DLEDuringTrialtarget", "prob_target_drt"),
  !!!ch_slot("NextBestMaxGainSamples", "DLEEndOfTrialtarget", "prob_target_eot"),
  !!!ch_slot("NextBestMaxGainSamples", "TDderive", "derive"),
  !!!ch_slot("NextBestMaxGainSamples", "Gstarderive", "mg_derive"),
  !!!ch_slot("IncrementsRelativeParts", "dltStart", "dlt_start", ""),
  !!!ch_slot("IncrementsRelativeParts", "cleanStart", "clean_start", "change applied also to other dependent classes"),
  !!!ch_slot(
    "IncrementsRelativeDLT", "DLTintervals", "dlt_intervals", "change applied also to other dependent classes"
  ),
  !!!ch_slot("IncrementsNumDoseLevels", "maxLevels", "max_levels", "change applied also to other dependent classes"),
  !!!ch_slot("IncrementsNumDoseLevels", "", "basis_level"),
  !!!ch_class("IncrementMin", "IncrementsMin"),
  !!!ch_slot("IncrementsMin", "IncrementsList", "increments_list", "change applied also to other dependent classes"),
  !!!ch_slot("StoppingTargetBiomarker", "scale", "", "replaced by `is_relative` flag"),
  !!!ch_slot("StoppingTargetBiomarker", "", "is_relative", "replaces `scale` string"),
  !!!ch_slot("StoppingList", "stopList", "stop_list"),
  !!!ch_slot("StoppingAll", "stopList", "stop_list"),
  !!!ch_slot("StoppingAny", "stopList", "stop_list"),
  !!!ch_slot("StoppingTDCIRatio", "targetRatio", "target_ratio"),
  !!!ch_slot("StoppingTDCIRatio", "targetEndOfTrial", "prob_target"),
  !!!ch_slot("StoppingGstarCIRatio", "targetRatio", "target_ratio"),
  !!!ch_slot("StoppingGstarCIRatio", "targetEndOfTrial", "prob_target"),
  !!!ch_class("StoppingGstarCIRatio", "StoppingMaxGainCIRatio"),
  !!!ch_slot("CohortSizeRange", "cohortSize", "cohort_size"),
)
```

```{r class_changes_kbl, echo=FALSE, results='markup'}

cc <- class_changes
cc <- cc[order(cc$class_name, cc$class_new_name, cc$slot_name, cc$slot_new_name), ]

no_slot_change <- nchar(trimws(cc$slot_name)) == 0
cc$class_name <- cell_spec(
  cc$class_name,
  strikeout = ifelse(cc$was_removed & no_slot_change, TRUE, FALSE)
)
cc$slot_name <- cell_spec(
  cc$slot_name,
  strikeout = ifelse(cc$was_removed & !no_slot_change, TRUE, FALSE)
)

cc <- cc[, -5]

kbl(cc, longtable = TRUE, booktabs = TRUE, col.names = NULL, escape = FALSE) %>%
  add_header_above(c("Name", "New Name", "Name", "New Name", "")) %>%
  add_header_above(c("Class/Constructor[note]" = 2, "Slot/Arg[note]" = 2, "Comment")) %>%
  column_spec(1:5, width = "6cm", border_left = TRUE, border_right = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 11) %>%
  add_footnote(
    c(
      "Class or class' user constructor. In the later case the Name/New Name is followed by ().",
      "Slot in case of the class or argument in case of the class' user constructor"
    ),
    notation = "symbol"
  )
```

Strikeout indicates that the class/slot was removed.

## Methods

```{r method_change_helpers, echo=FALSE}

# A simple helper function that generates change method record.
ch_method <- function(name = "", new_name = "", comment = "") {
  assert_string(name)
  assert_string(new_name)
  assert_string(comment)

  was_removed <- ifelse(nchar(new_name) == 0, TRUE, FALSE)
  list(name, new_name, "", "", was_removed, "", comment)
}

# A simple helper function that generates change argument record.
ch_arg <- function(method_name, name = "", new_name = "", comment = "") {
  assert_string(method_name)
  assert_string(name)
  assert_string(new_name)
  assert_string(comment)

  was_removed <- ifelse(nchar(new_name) == 0, TRUE, FALSE)
  list(method_name, "", name, new_name, was_removed, "", comment)
}

# A simple helper function that generates change output record.
ch_out <- function(method_name, output = "", comment = "") {
  checkmate::assert_string(method_name)
  checkmate::assert_string(output)
  checkmate::assert_string(comment)

  list(method_name, "", "", "", FALSE, output, comment)
}
```

```{r method_changes, echo=FALSE}

method_changes <- tribble(
  ~method_name, ~method_new_name, ~arg_name, ~arg_new_name, ~was_removed, ~output_changes, ~comment,
  !!!ch_arg("update-Data", "newCohort", "new_cohort"),
  !!!ch_arg("update-Data", "", "check"),
  !!!ch_arg("update-DataParts", "", "check"),
  !!!ch_arg("update-DataDual", "", "check"),
  !!!ch_arg("update-DataDA", "", "check"),
  !!!ch_out("getEff", "renamed elements in the list returned"),
  !!!ch_arg("mcmc", "fromPrior", "from_prior"),
  !!!ch_arg("mcmc", "program"),
  !!!ch_arg("mcmc", "verbose"),
  !!!ch_method("", "doseFunction", "returns dose computing function for a given model"),
  !!!ch_method("", "probFunction", "returns probability computing function for a given model"),
  !!!ch_arg("dose", "prob", "x"),
  !!!ch_method("", "efficacy", "new generic function"),
  !!!ch_method("biomLevel", "biomarker", "changed generic and methods"),
  !!!ch_arg("biomLevel", "dose", "", "changed generic and methods"),
  !!!ch_arg("biomLevel-DualEndpoint", "dose"),
  !!!ch_arg("biomLevel", "", "xLevel", "added only for generic, reordered arguments for generic and methods"),
  !!!ch_arg(
    "gain", "several arguments", "several arguments", "several arguments (generic and methods) changed, see man page"
  ),
  !!!ch_method("", "update-ModelPseudo"),
  !!!ch_method("update-LogisticIndepBeta", "", "replaced by `update-ModelPseudo`"),
  !!!ch_method("update-Effloglog", "", "replaced by `update-ModelPseudo`"),
  !!!ch_method("update-EffFlexi", "", "replaced by `update-ModelPseudo`"),
  !!!ch_arg("nextBest-NextBestTD", "Effmodel", "model_eff"),
  !!!ch_arg("nextBest-NextBestTD", "Effsamples", "samples_eff"),
  !!!ch_arg("nextBest-NextBestTD", "SIM", "in_sim"),
  !!!ch_arg("nextBest-NextBestTDsamples", "Effmodel", "model_eff"),
  !!!ch_arg("nextBest-NextBestTDsamples", "Effsamples", "samples_eff"),
  !!!ch_arg("nextBest-NextBestMaxGain", "Effmodel", "model_eff"),
  !!!ch_arg("nextBest-NextBestMaxGain", "Effsamples", "samples_eff"),
  !!!ch_arg("nextBest-NextBestMaxGain", "SIM", "in_sim"),
  !!!ch_out("nextBest-NextBestMaxGain", "renamed elements in the list returned"),
  !!!ch_arg("nextBest-NextBestMaxGainSamples", "Effmodel", "model_eff"),
  !!!ch_arg("nextBest-NextBestMaxGainSamples", "Effsamples", "samples_eff"),
  !!!ch_arg("nextBest-NextBestMaxGainSamples", "SIM", "in_sim"),
  !!!ch_out("nextBest-NextBestMaxGainSamples", "renamed elements in the list returned")
)
```

```{r method_changes_kbl, echo=FALSE, results='markup'}

mc <- method_changes
mc <- mc[order(mc$method_name, mc$method_new_name, mc$arg_name, mc$arg_new_name), ]

no_arg_change <- nchar(trimws(mc$arg_name)) == 0
mc$method_name <- cell_spec(
  mc$method_name,
  strikeout = ifelse(mc$was_removed & no_arg_change, TRUE, FALSE)
)
mc$arg_name <- cell_spec(
  mc$arg_name,
  strikeout = ifelse(mc$was_removed & !no_arg_change, TRUE, FALSE)
)
mc <- mc[, -5]

kbl(mc, longtable = TRUE, booktabs = TRUE, col.names = NULL, escape = FALSE) %>%
  add_header_above(c("Name", "New Name", "Name", "New Name", "", "")) %>%
  add_header_above(c("Method" = 2, "Argument" = 2, "Output Changes", "Comment")) %>%
  column_spec(1:6, width = "6cm", border_left = TRUE, border_right = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 11)
```

Strikeout indicates that the method/argument was removed.

# Moved `dose` and `prob` Functions from Slots to Methods

Moved `dose` and `prob` functions from model class slots to model class methods.
Example of usage: `dose`/`prob` function as a true dose-DLT/DLT-dose relationship.

## Generate data, define a model and get samples
```{r data_model_samples}
library(crmPack)

empty_data <- Data(doseGrid = c(1, 3, 5, 10, 15, 20, 25, 40, 50, 80, 100))
my_model <- LogisticNormal(
  mean = c(-0.85, 1),
  cov = matrix(c(1, -0.5, -0.5, 1), nrow = 2)
)
my_options <- McmcOptions(burnin = 2, step = 2, samples = 20)
my_samples <- mcmc(empty_data, my_model, my_options)
```

## Dose

Here is the example on how the `dose` function can be used in case of different
inputs, i.e. model's parameters samples or in case of a fixed model's parameters
values.

```{r dose_function}
# Doses reaching a specific target probability of the occurrence of a DLT (equal to 0.3),
# given a particular models and samples.
# Every single dose corresponds to one particular sample in `my_samples`.
dose(0.3, my_model, my_samples)

# True dose-DLT relationship.
# Say that -0.8 and 1 are the true values for models parameters alpha0 and alpha1 respectively.
# The `true_dose_fun` takes one argument (target probability of the occurrence of a DLT)
# and computes the corresponding dose, according to the model chosen and given a fixed values
# of the model's parameters.
true_dose_fun <- doseFunction(my_model, alpha0 = -0.8, alpha1 = 1)
true_dose_fun(0.3)
```

## Prob

```{r prob_function}
# Toxicity probabilities for a given dose (equal to 10), model and samples.
# Every single probability value corresponds to one particular sample in `my_samples`.
prob(10, my_model, my_samples)

# True DLT-dose relationship.
# Say that -0.8 and 1 are the true values for models parameters alpha0 and alpha1 respectively.
# The `true_prob_fun` takes one argument (the dose) and computes the corresponding
# toxicity probability, according to the model chosen and given a fixed values
# of the model's parameters.
true_prob_fun <- probFunction(my_model, alpha0 = -0.8, alpha1 = 1)
true_prob_fun(10)
```
