---
title: "Migrating from crmPack version 1.0.3"
author: "Wojciech Wójciak"
date: "05 October 2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from crmPack version 1.0.3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{crmPack}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(kableExtra)
library(checkmate)
```

The following vignette describes the changes that were introduced to the `crmPack`
package as a result of the package's refactoring. One row in below tables
represents a single-type, consistent change.

# Class and Methods Name Changes

## Classes

```{r class_changes, echo=FALSE}
class_changes <- tribble(
  ~class_name, ~class_new_name, ~slot_name, ~slot_new_name, ~was_removed, ~comment,
  "", "positive_number", "", "", FALSE, "to handle strictly positive valued slots
  (e.g. `ref_dose` in many model classes)",
  "", "ModelLogNormal", "", "", FALSE, "new parent class for all the models with reference dose and
  bivariate (log) normal prior on the model parameters",
  "", "ModelParamsNormal", "", "", FALSE, "represents parameters of bivariate normal distribution",
  "AllModels", "", "", "", TRUE, "",
  "Model", "", "", "", TRUE, "",
  "GeneralModel", "", "", "datanames", FALSE, "moved from `AllModels`",
  "GeneralModel", "", "", "datanames_prior", FALSE, "",
  "ModelEff", "", "dose", "", TRUE, "moved to model class method",
  "ModelEff", "", "ExpEff", "", TRUE, "moved to model class method and renamed to `efficacy`",
  "Effloglog", "", "several slots", "several slots", FALSE, "several slot changed, see man page",
  "EffFlexi", "", "several slots", "several slots", FALSE, "several slot changed, see man page",
  "EffFlexi()", "", "smooth", "", TRUE, "replaced by `rw1` flag",
  "EffFlexi()", "", "", "rw1", FALSE, "replaces `smooth` string",
  "DALogisticLogNormal", "", "C_par", "c_par", FALSE, "",
  "DALogisticLogNormal", "", "conditionalPEM", "cond_pem", FALSE, "",
  "TITELogisticLogNormal", "", "weightMethod", "weight_method", FALSE, "",
  "OneParLogNormalPrior", "", "skeletonFun", "skel_fun", FALSE, "",
  "OneParLogNormalPrior", "", "", "skel_fun_inv", FALSE, "",
  "OneParLogNormalPrior", "", "skeletonProbs", "skel_probs", FALSE, "",
  "OneParLogNormalPrior()", "", "doseGrid", "dose_grid", FALSE, "",
  "DualEndpointBeta", "", "refDoseBeta", "ref_dose_beta", FALSE, "",
  "DualEndpointEmax", "", "refDoseEmax", "ref_dose_emax", FALSE, "",
  "DualEndpointRW", "", "useRW1", "rw1", FALSE, "",
  "DualEndpointRW()", "", "smooth", "", TRUE, "replaced by `rw1` flag",
  "DualEndpointRW()", "", "", "rw1", FALSE, "replaces `smooth` string",
  "McmcOptions", "", "", "rng_kind", FALSE, "to be used by Random Number Generator in rJAGS",
  "McmcOptions", "", "", "rng_seed", FALSE, "to be used by Random Number Generator in rJAGS",
  "NextBestTD", "", "targetDuringTrial", "prob_target_drt", FALSE, "",
  "NextBestTD", "", "targetEndOfTrial", "prob_target_eot", FALSE, "",
  "NextBestTDsamples", "", "targetDuringTrial", "prob_target_drt", FALSE, "",
  "NextBestTDsamples", "", "targetEndOfTrial", "prob_target_eot", FALSE, "",
  "NextBestNCRM", "", "maxOverdoseProb", "max_overdose_prob", FALSE, "",
  "NextBestDualEndpoint", "", "scale", "", TRUE, "replaced by `target_relative` flag",
  "NextBestDualEndpoint", "", "", "target_relative", FALSE, "replaces `scale` string",
  "NextBestDualEndpoint", "", "maxOverdoseProb", "max_overdose_prob", FALSE, "",
  "NextBestDualEndpoint", "", "targetThresh", "target_thresh", FALSE, "",
  "NextBestMaxGain", "", "DLEDuringTrialtarget", "prob_target_drt", FALSE, "",
  "NextBestMaxGain", "", "DLEEndOfTrialtarget", "prob_target_eot", FALSE, "",
  "NextBestMaxGainSamples", "", "DLEDuringTrialtarget", "prob_target_drt", FALSE, "",
  "NextBestMaxGainSamples", "", "DLEEndOfTrialtarget", "prob_target_eot", FALSE, "",
  "NextBestMaxGainSamples", "", "TDderive", "derive", FALSE, "",
  "NextBestMaxGainSamples", "", "Gstarderive", "mg_derive", FALSE, "",
  "IncrementsRelativeParts", "", "dltStart", "dlt_start", FALSE, "",
  "IncrementsRelativeParts", "", "cleanStart", "clean_start", FALSE, "change applied also to other dependent classes",
  "IncrementsRelativeDLT", "", "DLTintervals", "dlt_intervals", FALSE, "change applied also to other dependent classes",
  "IncrementsNumDoseLevels", "IncrementsDoseLevels", "", "", FALSE, "",
  "IncrementsDoseLevels", "", "maxLevels", "levels", FALSE, "change applied also to other dependent classes",
  "IncrementsDoseLevels", "", "", "basis_level", FALSE, "",
  "IncrementMin", "IncrementsMin", "", "", FALSE, "",
  "IncrementsMin", "", "IncrementsList", "increments_list", FALSE, "change applied also to other dependent classes",
  "StoppingTargetBiomarker", "", "scale", "", TRUE, "replaced by `is_relative` flag",
  "StoppingTargetBiomarker", "", "", "is_relative", FALSE, "replaces `scale` string",
  "StoppingList", "", "stopList", "stop_list", FALSE, "",
  "StoppingAll", "", "stopList", "stop_list", FALSE, "",
  "StoppingAny", "", "stopList", "stop_list", FALSE, "",
  "StoppingTDCIRatio", "", "targetRatio", "target_ratio", FALSE, "",
  "StoppingTDCIRatio", "", "targetEndOfTrial", "prob_target", FALSE, "",
  "StoppingGstarCIRatio", "", "targetRatio", "target_ratio", FALSE, "",
  "StoppingGstarCIRatio", "", "targetEndOfTrial", "prob_target", FALSE, "",
  "StoppingGstarCIRatio", "StoppingMaxGainCIRatio", "", "", FALSE, "",
  "CohortSizeRange", "", "cohortSize", "cohort_size", FALSE, "",
  "CohortSizeDLT", "", "DLTintervals", "dlt_intervals", FALSE, "",
  "CohortSizeDLT", "", "cohortSize", "cohort_size", FALSE, "",
  "CohortSizeMax", "", "cohortSizeList", "cohort_size_list", FALSE, "",
  "CohortSizeMin", "", "cohortSizeList", "cohort_size_list", FALSE, "",
  "SafetyWindowSize", "", "patientGap", "gap", FALSE, "",
  "SafetyWindowSize", "", "sizeIntervals", "size", FALSE, "",
  "SafetyWindowSize", "", "patientFollow", "follow", FALSE, "",
  "SafetyWindowSize", "", "patientFollowMin", "follow_min", FALSE, "",
  "SafetyWindowConst", "", "patientGap", "gap", FALSE, "",
  "SafetyWindowConst", "", "patientFollow", "follow", FALSE, "",
  "SafetyWindowConst", "", "patientFollowMin", "follow_min", FALSE, "",
  "OneParExpNormalPrior", "OneParLogNormalPrior", "", "", FALSE, "",
  "Design", "", "PLcohortSize", "pl_cohort_size", FALSE, "",
  "TDsamplesDesign", "", "PLcohortSize", "pl_cohort_size", FALSE, "",
  "TDDesign", "", "PLcohortSize", "pl_cohort_size", FALSE, ""
)
```

```{r class_changes_kbl, echo=FALSE, results='markup'}
cc <- class_changes
cc <- cc[order(cc$class_name, cc$class_new_name, cc$slot_name, cc$slot_new_name), ]

no_slot_change <- nchar(trimws(cc$slot_name)) == 0
cc$class_name <- cell_spec(
  cc$class_name,
  strikeout = ifelse(cc$was_removed & no_slot_change, TRUE, FALSE)
)
cc$slot_name <- cell_spec(
  cc$slot_name,
  strikeout = ifelse(cc$was_removed & !no_slot_change, TRUE, FALSE)
)

cc <- cc[, -5]

kbl(cc, longtable = TRUE, booktabs = TRUE, col.names = NULL, escape = FALSE) %>%
  add_header_above(c("Name", "New Name", "Name", "New Name", "")) %>%
  add_header_above(c("Class/Constructor[note]" = 2, "Slot/Arg[note]" = 2, "Comment")) %>%
  column_spec(1:5, width = "6cm", border_left = TRUE, border_right = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 11) %>%
  add_footnote(
    c(
      "Class or class' user constructor. In the later case the Name/New Name is followed by ().",
      "Slot in case of the class or argument in case of the class' user constructor"
    ),
    notation = "symbol"
  )
```

Strikeout indicates that the class/slot was removed.

## Methods

```{r method_changes, echo=FALSE}
method_changes <- tribble(
  ~method_name, ~method_new_name, ~arg_name, ~arg_new_name, ~was_removed, ~output_changes, ~comment,
  "update-Data", "", "newCohort", "new_cohort", FALSE, "", "",
  "update-Data", "", "", "check", FALSE, "", "",
  "update-DataParts", "", "", "check", FALSE, "", "",
  "update-DataDual", "", "", "check", FALSE, "", "",
  "update-DataDA", "", "", "check", FALSE, "", "",
  "getEff", "", "", "", FALSE, "renamed elements in the list returned", "",
  "mcmc", "", "fromPrior", "from_prior", FALSE, "", "",
  "mcmc", "", "program", "", TRUE, "", "",
  "mcmc", "", "verbose", "", TRUE, "", "",
  "", "doseFunction", "", "", FALSE, "", "returns dose computing function for a given model",
  "", "probFunction", "", "", FALSE, "", "returns probability computing function for a given model",
  "dose", "", "prob", "x", FALSE, "", "",
  "", "efficacy", "", "", FALSE, "", "new generic function",
  "biomLevel", "biomarker", "", "", FALSE, "", "changed generic and methods",
  "biomLevel", "", "dose", "", TRUE, "", "changed generic and methods",
  "biomLevel-DualEndpoint", "", "dose", "", TRUE, "", "",
  "biomLevel", "", "", "xLevel", FALSE, "", "added only for generic, reordered arguments for generic and methods",
  "gain", "", "several arguments", "several arguments", FALSE, "", "several arguments (generic and methods) changed,
  see man page",
  "", "update-ModelPseudo", "", "", FALSE, "", "",
  "update-LogisticIndepBeta", "", "", "", TRUE, "", "replaced by `update-ModelPseudo`",
  "update-Effloglog", "", "", "", TRUE, "", "replaced by `update-ModelPseudo`",
  "update-EffFlexi", "", "", "", TRUE, "", "replaced by `update-ModelPseudo`",
  "nextBest-NextBestTD", "", "Effmodel", "model_eff", FALSE, "", "",
  "nextBest-NextBestTD", "", "Effsamples", "samples_eff", FALSE, "", "",
  "nextBest-NextBestTD", "", "SIM", "in_sim", FALSE, "", "",
  "nextBest-NextBestTDsamples", "", "Effmodel", "model_eff", FALSE, "", "",
  "nextBest-NextBestTDsamples", "", "Effsamples", "samples_eff", FALSE, "", "",
  "nextBest-NextBestMaxGain", "", "Effmodel", "model_eff", FALSE, "", "",
  "nextBest-NextBestMaxGain", "", "Effsamples", "samples_eff", FALSE, "", "",
  "nextBest-NextBestMaxGain", "", "SIM", "in_sim", FALSE, "", "",
  "nextBest-NextBestMaxGain", "", "", "", FALSE, "renamed elements in the list returned", "",
  "nextBest-NextBestMaxGainSamples", "", "Effmodel", "model_eff", FALSE, "", "",
  "nextBest-NextBestMaxGainSamples", "", "Effsamples", "samples_eff", FALSE, "", "",
  "nextBest-NextBestMaxGainSamples", "", "SIM", "in_sim", FALSE, "", "",
  "nextBest-NextBestMaxGainSamples", "", "", "", FALSE, "renamed elements in the list returned", "",
  "size", "size", "", "check", FALSE, "", ""
)
```

```{r method_changes_kbl, echo=FALSE, results='markup'}
mc <- method_changes
mc <- mc[order(mc$method_name, mc$method_new_name, mc$arg_name, mc$arg_new_name), ]

no_arg_change <- nchar(trimws(mc$arg_name)) == 0
mc$method_name <- cell_spec(
  mc$method_name,
  strikeout = ifelse(mc$was_removed & no_arg_change, TRUE, FALSE)
)
mc$arg_name <- cell_spec(
  mc$arg_name,
  strikeout = ifelse(mc$was_removed & !no_arg_change, TRUE, FALSE)
)
mc <- mc[, -5]

kbl(mc, longtable = TRUE, booktabs = TRUE, col.names = NULL, escape = FALSE) %>%
  add_header_above(c("Name", "New Name", "Name", "New Name", "", "")) %>%
  add_header_above(c("Method" = 2, "Argument" = 2, "Output Changes", "Comment")) %>%
  column_spec(1:6, width = "6cm", border_left = TRUE, border_right = TRUE) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  kable_styling(latex_options = c("repeat_header"), font_size = 11)
```

Strikeout indicates that the method/argument was removed.

# Moved `dose` and `prob` Functions from Slots to Methods

Moved `dose` and `prob` functions from model class slots to model class methods.
Example of usage: `dose`/`prob` function as a true dose-DLT/DLT-dose relationship.

## Generate data, define a model and get samples
```{r data_model_samples}
library(crmPack)

empty_data <- Data(doseGrid = c(1, 3, 5, 10, 15, 20, 25, 40, 50, 80, 100))
my_model <- LogisticNormal(
  mean = c(-0.85, 1),
  cov = matrix(c(1, -0.5, -0.5, 1), nrow = 2)
)
my_options <- McmcOptions(burnin = 2, step = 2, samples = 20)
my_samples <- mcmc(empty_data, my_model, my_options)
```

## Dose

Here is the example on how the `dose` function can be used in case of different
inputs, i.e. model's parameters samples or in case of a fixed model's parameters
values.

```{r dose_function}
# Doses reaching a specific target probability of the occurrence of a DLT (equal to 0.3),
# given a particular models and samples.
# Every single dose corresponds to one particular sample in `my_samples`.
dose(0.3, my_model, my_samples)

# True dose-DLT relationship.
# Say that -0.8 and 1 are the true values for models parameters alpha0 and alpha1 respectively.
# The `true_dose_fun` takes one argument (target probability of the occurrence of a DLT)
# and computes the corresponding dose, according to the model chosen and given a fixed values
# of the model's parameters.
true_dose_fun <- doseFunction(my_model, alpha0 = -0.8, alpha1 = 1)
true_dose_fun(0.3)
```

## Prob

```{r prob_function}
# Toxicity probabilities for a given dose (equal to 10), model and samples.
# Every single probability value corresponds to one particular sample in `my_samples`.
prob(10, my_model, my_samples)

# True DLT-dose relationship.
# Say that -0.8 and 1 are the true values for models parameters alpha0 and alpha1 respectively.
# The `true_prob_fun` takes one argument (the dose) and computes the corresponding
# toxicity probability, according to the model chosen and given a fixed values
# of the model's parameters.
true_prob_fun <- probFunction(my_model, alpha0 = -0.8, alpha1 = 1)
true_prob_fun(10)
```

# New Random Number Generator settings for the MCMC

The Random Number Generator (RNG) settings used by the JAGS for the MCMC are now
configured solely through the `McmcOptions` class. The RNG settings are:
`RNG type` and the `RNG seed` that corresponds to a given `RNG type`. Find out
details in the help page for the `McmcOptions` class.
Any RNG-related user settings at the R session level (such us those with
`set.seed()`) are ignored by the MCMC sampler.

