---
title: "Using tidy methods"
author: "John Kirkpatrick"
date: "20Jul2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(broom)
devtools::load_all()
```

## Introduction
Here is a first draft implementation of broom-like methods for `crmPack, as suggested in [issue 323](https://github.com/Roche/crmPack/issues/323).

> The original version of this specification did not account for slots whose values were lists.  This update corrects that omission.

Whilst variants of "How do I convert an S4 object to a `data.frame`/`tibble`?" are commonly asked questions on StackOverflow [see, for example, [here](https://stackoverflow.com/questions/57317958/general-way-to-transform-s4-object-to-dataframe-r) and [here](https://stackoverflow.com/questions/14604269/how-to-create-a-dataframe-of-user-defined-s4-classes-in-r)], there appears to be no existing package to do this.  However, the approach generally suggested in answers to these questions is essentially the same as suggested here: 

*  Deal with the object slot by slot  
*  Convert each slot to a `tibble`  
*  Return the slot-specific `tibble`s in a `list` 

### Principles
The general principles I propose for tidying `crmPack` objects are as follows:

* All slots that are not functions are converted to `tibble`s or a list of `tibble`s.
* If the slot's value is scalar, the slot is converted to a 1x1 `tibble`.  This will ease downstream operations such as `row_bind`ing.
* If the slot's value is a `list`, these rules are aplied to each element of the list in turn.
* If the object being tidied contains multiple slots of (potentially) different lengths, the results is a list of `tibble`s.  The list may be nested to multiple levels.  (See, for example, `LogisticLogNormal`.)
* The column names of the tidied `tibble` correspond to the slot names of the parent object.
  * Exception: where the slot has name in the plural and contains a `vector` or `list`, the column name will be singular.  See, for example, `CohortSizeParts` below.
* When the value of a slot has not been set, a zero-row `tibble` is returned.
* When the value of a slot has scalar attributes, these attributes are added as a column of the `tibble`, whose name is the name of the attribute and whose value is the value of the attribute for every row of the tibble.  Vector attributes can be added, by default, as a nested tibble.
  * Two formats for the nested tibble are possible.  First, 1 row x n column, with column names defined by the name of the attribute and values given by the value of the corresponding attribute.  Second, an n row by 2 column tibble, with columns (say) `name` and `value`.  If different attributes have different value types, then the `value` column would most likely be coerced to character.  For this reason, I prefer the first option.
* `tbl_<className>` is prepended to the class of the (list of) tidy `tibble`(s).
  
[Issue 652](https://github.com/Roche/crmPack/issues/652) has introduced a common super class for all other `crmPack` classes.  This can be leveraged to provide a single `tidy()` method that will tidy the majority of class-to-tibble conversions in `crmPack`.  A helper function to manage interval conversions (as discussed under Exceptions below) would allow almost every exception to the general rule to be handled in a simple, two line sub class-specific method.  

### Exceptions
*  Where a vector slot (or series of vector slots) define a range, then the naming convention described above is not followed.  Instead, columns named `min` and `max` define the extent of the range and a column named `Range` labels the range.
  * Would, say, `left` and `right` be better names to avoid potential name clashes with the built-in `min` and `max` functions? 


### Helper functions
It is likely that `tidy`ing `crmPack` objects of different classes will share common steps.  These common steps are candidates for being handled by helper functions.  Functionality for various helper functions include:

* Conversion of a single slot to a tibble [`h_tidy_slot` with signature `c(obj = ANY, slot_name = "character")`]
* Conversion of all slots in an object to tibbles [`h_tidy_all_slots` with signature `c(obj = ANY)`] that simply calls `h_tidy_slot` for all slots in the object that are not functions.
* The conversion of scalar attributes to "constant" columns in a tibble [`h_scalar_attr_to_column` with signature `c(data = "tibble", obj = ANY)`]
* The conversion of ranges (as found in, for example, sub-classes `Increments`) to min/max column pairs [`h_range_to_minmax` with signature `c(data = "tibble", obj = ANY)` with additional arguments to allow users to specify the name of the min and max columns, whether the ranges are closed to the left (default) or right and the outer ends of the lowest and highest intervals (with defaults `-Inf` and `Inf` respectively)]
* The handling of class attributes
* The class definition of the `tidy` tibble [`h_tidy_class` with signature `c(data = "tibble")`]

### Please also read
Please also read [Issue 407](https://github.com/Roche/crmPack/issues/407) regarding inconsistent slot names.

## Implementation
`tidy()` is a generic defined in `broom`.  An end-user of `crmPack` may not have installed `broom`.  Therefore, we have two options:

* Make `crmPack` Depend on `broom`
* On loading `crmPack`, determine if `broom` is an available namespace and, if not, define the appropriate generic

After discussion, we take the first approach and make `crmPack` Depend on `broom`.

### Some helper functions

To handle the attributes:

```{r}
h_handle_attributes <- function(x, .ignore = c("names", "class", "description", "row.names")) {
  a <- attributes(x)
  valid_names <- setdiff(names(a), .ignore)
  lapply(
    valid_names,
    function(n) {
      z <- attr(x, n)
      if (length(z) == 1) {
        rv <- tibble::tibble(X = z)
      } else {
        rv <- tibble::tibble(X = list(z))
      }
      names(rv) <- n
      rv
    }
  ) |>
    dplyr::bind_cols()
}
```

Other helpers.

```{r define-helper-functions}
h_tidy_slot <- function(obj, slot_name, col = NULL, attributes = FALSE) {
  if (is.list(slot(obj, slot_name))) {
    return(lapply(slot(obj, slot_name), tidy))
  }
  a <- h_handle_attributes(slot(obj, slot_name))
  if (is(slot(obj, slot_name), "CrmPackClass")) {
    rv <- slot(obj, slot_name) |>
      tidy() |>
      dplyr::bind_cols()
  } else {
    if (is.null(col)) {
      col <- slot_name
    }
    rv <- tibble::tibble({{ col }} := slot(obj, slot_name))
  }
  if (nrow(a) > 0 & attributes) {
    # rv <- rv |> dplyr::bind_cols(a)
    print(slot_name)
    print(a)
  }
  rv
}

h_tidy_all_slots <- function(obj) {
  slot_names <- slotNames(obj)
  rv <- list()
  for (nm in slot_names) {
    if (!is.function(slot(obj, nm))) {
      rv[[nm]] <- h_tidy_slot(obj, nm)
    }
  }
  # Column bind of all list elements have the same number of rows
  if (length(rv) > 1 & length(unique(sapply(rv, nrow))) == 1) {
    rv <- rv |> dplyr::bind_cols()
  } 
  rv
}

h_tidy_class <- function(d, obj) {
  cls <- class(obj)
  class(d) <- c(paste0("tbl_", cls[1]), class(d))
  d
}

h_range_to_minmax <- function(
    x,
    col,
    min_col = "min",
    max_col = "max",
    range_min = -Inf,
    range_max = Inf) {
  vals <- x %>% dplyr::pull({{ col }})
  tibble(
    {{ min_col }} := c(range_min, vals),
    {{ max_col }} := c(vals, range_max)
  )
}
```

### The default `tidy` function

```{r}
#' @param x (`CrmPackClass`)\cr The objecte to be tidied
#' @param ... \cr Not used at present
#' @rdname tidy
#' @aliases tidy-CrmCPackClass
#' @example examples/CrmPackClass-method-tidy.R
#' @export
setMethod(
  f = "tidy",
  signature = signature(x = "CrmPackClass"),
  definition = function(x, ...) {
    rv <- h_tidy_all_slots(x) |> h_tidy_class(x)
    if (length(rv) == 1) {
      rv[[names(rv)[1]]]
    } else {
      rv
    }
  }
)
```

### Simple examples

`CohortSizeConst` is a trivial example and illustrates the default approach for all classes.

```{r, error = TRUE}
x <- CohortSizeConst(size = 3) |> tidy()
x
class(x)
```

`IncrementsRelative` illustrate how ranges are handled. Each class that includes a range will require individual handling because there's no generic way to identify which slot defines the range, nor how it should be handled.

```{r, error = TRUE}
# tidy-IncrementsRelative ----

#' @rdname tidy
#' @aliases tidy-IncrementsRelative
#' @example examples/IncrementsRelative-method-tidy.R
#' @export
setMethod(
  f = "tidy",
  signature = signature(x = "IncrementsRelative"),
  definition = function(x, ...) {
    h_tidy_all_slots(x) |>
      dplyr::bind_cols() |>
      h_range_to_minmax(intervals) |>
      dplyr::filter(max > 0) |>
      tibble::add_column(increment = x@increments) |>
      h_tidy_class(x)
  }
)

x <- IncrementsRelative(
  intervals = c(0, 20),
  increments = c(1, 0.33)
) |>
  tidy()

x
class(x)
```

`CohortSizeMax` contains a slot whose value is a list.

```{r, error = TRUE}
cs_max <- maxSize(
  CohortSizeConst(3), 
  CohortSizeDLT(intervals = 0:1, cohort_size = c(1, 3))
)
cs_max |> tidy()
```

`CohortSizeParts` is slightly more complex and requires custom handling because the part index is implicit rather than explicit.

```{r}
# tidy-CohortSizeParts ----

#' @rdname tidy
#' @aliases tidy-CohortSizeParts
#' @example examples/CohortSizeParts-method-tidy.R
#' @export
setMethod(
  f = "tidy",
  signature = signature(x = "CohortSizeParts"),
  definition = function(x, ...) {
    tibble::tibble(
      part = seq_along(x@cohort_sizes),
      cohort_size = x@cohort_sizes
    ) |>
      h_tidy_class(x)
  }
)

x <- CohortSizeParts(cohort_sizes = c(1, 3)) |> tidy()
x
class(x)
```

`NextBestNCRM` also requires custom handling because the `overdose_prob` slot applies only to the "Overdose" element of the range.  

```{r, error = TRUE}
# tidy-NextBestNCRM ----

#' @rdname tidy
#' @aliases tidy-NextBestNCRM
#' @example examples/NextBestNCRM-method-tidy.R
#' @export
setMethod(
  f = "tidy",
  signature = signature(x = "NextBestNCRM"),
  definition = function(x, ...) {
    rv <- h_tidy_all_slots(x) |>
      dplyr::bind_cols() |>
      h_range_to_minmax(target, range_min = 0, range_max = 1) |>
      add_column(max_prob = c(NA, NA, x@max_overdose_prob)) |>
      add_column(Range = c("Underdose", "Target", "Overdose"), .before = 1)
  }
)

x <- NextBestNCRM(
  target = c(0.2, 0.35),
  overdose = c(0.35, 1),
  max_overdose_prob = 0.25
) |> tidy()
x
class(x)
```
In addition, consideration of how to handle the case when the lower end of `overdose` is above the upper end of `target` because `crmPack` does not follow the usual convention of classifying p(Tox) into four categories, namely, "Underdosing", "Target", "Overdosing" and "Toxicity".  This implementation gives incorrect output:

```{r, error = TRUE}
NextBestNCRM(
  target = c(0.2, 0.35),
  overdose = c(0.6, 1),
  max_overdose_prob = 0.25
) |> tidy()
```

Should the output be similar to

```{r echo = FALSE}
tibble(
  Range = c("Underdose", "Target", "Overdose"),
  min = c(0, 0.2, 0.6),
  max = c(0.2, 0.35, 1),
  max_prob = c(NA, NA, 0.25)
)
```

or similar to 

```{r echo = FALSE}
tibble(
  Range = c("Underdose", "Target", " ", "Overdose"),
  min = c(0, 0.2, 0.35, 0.6),
  max = c(0.2, 0.35, 0.6, 1),
  max_prob = c(NA, NA, NA, 0.25)
)
```
?

### More complex examples

Various sub-classes of `GeneralModel` demonstrate how `tidy()` handles both slots of different classes and nesting.  Here is `LogisticLogNormal` as an example.

```{r, error=TRUE}
x <- LogisticLogNormal(
  mean = c(-0.85, 1),
  cov = matrix(c(1, -0.5, -0.5, 1), nrow = 2),
  ref_dose = 50
) 

x |> tidy()
```

> The `cov` and `prec` slots of the `ModelParamsLogNormal` class both have an attribute set.  Unfortunately, it's called `dim` in both cases.  As might be expected, the value of `dim` is the same for both slots.  As `dim` is uninteresting, this can be easily dealt with: by ignoring it.

The `McmcOptions` used to create a `Samples` object are added as attributes to each member of the `data` slot.  So this would also benefit from custom handling.

```{r}
setMethod(
  f = "tidy",
  signature = signature(x = "Samples"),
  definition = function(x, ...) {
    rv <- lapply(
      slotNames(x),
      function(nm) {
        if (nm == "data") {
          lapply(
            names(x@data),
            function(nm) {
              as_tibble(get(x, nm))
            }
          ) |>
            dplyr::bind_rows() |>
            tidyr::pivot_wider(
              names_from = Parameter,
              values_from = value
            ) |>
            bind_cols(h_handle_attributes(get(x, names(x@data)[1])))
        } else {
          slot(x, nm) |>
            tidy() |>
            dplyr::bind_cols()
        }
      }
    )
    names(rv) <- c("data", "options")
    rv <- rv |> h_tidy_class(x)
    rv
  }
)

options <- McmcOptions(
  burnin = 100,
  step = 1,
  samples = 2000
)

emptydata <- Data(doseGrid = c(1, 3, 5, 10, 15, 20, 25, 40, 50, 80, 100))

model <- LogisticLogNormal(
  mean = c(-0.85, 1),
  cov =
    matrix(c(1, -0.5, -0.5, 1),
      nrow = 2
    ),
  ref_dose = 56
)
x <- mcmc(emptydata, model, options)
y <- x %>% tidy()
class(y)
```

> `ModelParamsNormal` and `Samples` are the only two class I've found that set attributes.  Both need special handling, so I've not been able to demonstrate "default" handling of parameters.  Are there any other classes that set attributes?

Since the `options` tibble is a one row copy of data contained in every row of the `data` tibble, it's value is questionable.  There is a case for returning only the `data` element of the default list.

## A useful side effect

By introducing the `tidy` method, we have an immediate benefit of greatly improving the presentation of `crmPack` objects in reports.  For example:

```{r, error = TRUE}
y[["data"]] |>
  head(20) |>
  kable(
    digits = c(0, 0, 4, 4, 0, 0, 0, 0, 0, NA),
    table.attr = "style = 'width: 100%;'"
  )
```

and

```{r}
NextBestNCRM(
  target = c(0.2, 0.35),
  overdose = c(0.6, 1),
  max_overdose_prob = 0.25
) |>
  tidy() |>
  kable(table.attr = "style = 'width: 40%;'")
```

Though whether we should introduce an additional dependency on `knitr` is moot.

## Default tidying for all crmPack classes

Errors indicate classes that will require custom processing, or for which tidying is inappropriate.

```{r, error = TRUE}
crmPack_class_list <- getClasses(asNamespace("crmPack"))

for (cls in crmPack_class_list) {
  if (!isClassUnion(cls)) {
    constructor_name <- paste0(".Default", cls)
    if (exists(constructor_name, mode = "function")) {
      print(cls)
      tryCatch({
        x <- do.call(paste0(".Default", cls), list())
        print(x |> tidy())
      },
      error = function(e) print(paste0("Unable to tidy ", cls, " objects."))
      )
    } else {
      print(paste0("No default constructor for ", cls))
    }
  }
}
```
