---
title: "Design for Two Drugs Combination"
author: "Daniel"
date: today
format: html
embed-resources: true
bibliography: references.bib
---

## Motivation

In the context of dose-finding for drug combinations, we are interested in modeling the probability of dose-limiting toxicity (DLT) as a function of the doses of two drugs. The goal is to identify the optimal combination of doses that maximizes efficacy while minimizing toxicity.

For the time being, we restrict the discussion to the case of two drugs, but the approach can be extended to more drugs in principle. However this seems to be not so relevant in practice, and could later extend the implementation to more drugs if needed.

## Idea

The idea is to implement the 5-parameter BLRM approach by @Neuenschwander2015 (note that a presentation slide deck is available [here](https://de.slideshare.net/slideshow/eugm-2014-roychaudhuri-phase-1-combination/50032570)):

- Let $\textrm{odds}(p)=p/(1-p)$ be the odds transformation of the probability $p$, such that 
  $\textrm{logit}(p) = \log(\textrm{odds}(p))$. 
- Let $x_i$ be the dose of drug $i=1,2$, and $p(x_1, x_2)$ be the probability of DLT with doses $x_1$ and $x_2$.
- The reference doses for the two compounds are again denoted by $x_1^{*}$ and $x_2^{*}$, and the single-agent DLT probabilities at these reference doses are denoted by $p(x_1)$ and $p(x_2)$, respectively.
- Then the model assumes a linear interaction function:

  $$
  \textrm{odds}(p(x_1, x_2)) = \textrm{odds}(p_0(x_1, x_2)) \cdot \exp(\eta x_1/x_1^{*} x_2/x_2^{*}),
  $$

  - where $\eta$ is the interaction coefficient (positive values correspond to synergistic toxicity, zero corresponds to additive effect without interaction, and negative values correspond to antagonistic toxicity). 
  - This model is also called the "dual combo model" by Novartis
  - Under no interaction with $\eta=0$, this reduces the probability $\textrm{odds}(p(x_1, x_2))$ to
  \begin{align*}
  p_0(x_1, x_2) &= p(x_1) + p(x_2) - p(x_1)p(x_2)\\
   &= 1 - (1 - p(x_1))(1 - p(x_2)).
  \end{align*}

- Now for the single-agent DLT probabilities $p(x_1)$ and $p(x_2)$ we assume the logistic log-normal models:
  $\textrm{logit}[p(x_i)] = \alpha_i + \beta_i \log(x_i/x_i^{*}),$
  with prior e.g.
  $(\alpha_i, \log(\beta_i))^\top \sim \textrm{Normal}(\mu_i, \Sigma_i)$
  for $i=1,2$. 

### Priors

- Different priors can be used in principle for the 5 parameters $\theta = (\alpha_1, \beta_1, \alpha_2, \beta_2, \eta)^\top$.
- Typical is a meta-analytic combined (MAC) prior which separates according to different trial sources:
  - mono drug 1 trial ($\theta_1$), which will only inform the parameters $\alpha_1$ and $\beta_1$ for drug 1
  - current combination of drugs 1 and 2 trial ($\theta_2$), which will inform all 5 parameters
  - mono drug 2 trial ($\theta_3$), which will only inform the parameters $\alpha_2$ and $\beta_2$ for drug 2
- And then uses the hierarchical prior $\theta_j \sim \textrm{Normal}(\mu, \Sigma)$, $j=1,2,3$.
- Hyperpriors on $\mu = (\mu_{\alpha_1}, \mu_{\beta_1}, \mu_{\alpha_2}, \mu_{\beta_2}, \mu_{\eta})^\top$ e.g.:
  - $\mu_{\alpha_i} \sim \textrm{Normal}(\textrm{logit}(0.25), 2.5^2)$, $i=1,2$
  - $\mu_{\beta_1} \sim \textrm{Normal}(0, 0.7^2)$
  - $\mu_{\beta_2}, \mu_{\eta} \sim \textrm{Normal}(0, 1)$
- Structured assumptions for $\Sigma$
  - Assume that only $\mu_{\alpha_1}$ and $\mu_{\beta_1}$, as well as $\mu_{\alpha_2}$ and $\mu_{\beta_2}$ are correlated, but not the other parameters, then we have a corresponding block diagonal structure for $\Sigma$ with 7 parameters
  (standard deviations $\tau_{\alpha_1}, \tau_{\beta_1}, \tau_{\alpha_2}, \tau_{\beta_2}, \tau_{\eta}$ for each of the 5 parameters plus the two correlations $\rho_1$ and $\rho_2$ between $\alpha_i$ and $\beta_i$ for $i=1,2$).
  - We don't fix the $\Sigma$ entries but instead again assign hyperpriors, e.g. with $\kappa = \log(2)/1.96$
    - $\tau_{\alpha_1} \sim \textrm{LogNormal}(\log(0.5), \kappa^2)$
    - $\tau_{\alpha_2} \sim \textrm{LogNormal}(\log(0.75), \kappa^2)$
    - $\tau_{\beta_1}, \tau_{\beta_2} \sim \textrm{LogNormal}(\log(0.25), \kappa^2)$
    - $\tau_{\eta} \sim \textrm{LogNormal}(\log(0.125), \kappa^2)$
    - $\rho_1, \rho_2 \sim \textrm{Uniform}(-1, 1)$

### Design

In terms of possible trial designs, we need to support the following options:

- Combination dose escalation with a fixed prior
- Combination dose escalation with historical data informing the prior for the monotherapy parameters
- Combination dose escalation with concurrent monotherapy arm or arms

The idea of these trial designs is that it must be possible to "plug together" multiple arms (mono drug 1, mono drug 2, combo drug 1 and 2, historical drug 1, historical drug 2, etc.) in a sort of "hierarchical design" object. Each arm has its own data, its own recommendation, stopping and increment rules, its own backfilling choices, and is simulated independently and might also start at different times, but parts of the models across the arms can be shared (e.g. as described above for the 5-parameters BLRM model).

One more detail here is that it should also be possible that the combo arm uses the mono data, but not the other way around necessarily - so it should be possible to avoid information flowing from the combo arm to the mono arm. On the other hand, this is optional, so in another choice the information could also flow from the combo arm to the mono arm, which would be more efficient in terms of learning about the single-agent parameters.

In most cases one of the combo drugs will have a fixed dose (e.g. the standard of care), and the other drug will be escalated. In the remaining cases, both drugs will be escalated. In that case, we need to define the recommendation rule: For the beginning it would be sufficient to extend the existing rule that chooses the dose with the largest target toxicity probability among the admissible doses, to the case of dose combinations.

Backfilling is often only interesting for the combo arm, but not for the mono arms. So backfilling needs to be configurable at the arm level. The same applies to the maximum increment rules and stopping rules.

### Increment rules

Increment rules can be defined at the arm level. Here for the combo arm we will need to extend the existing rules to the case of dose combinations. For example, we can define a maximum increment rule that allows only dose increments of one level for each drug, or even just for one of the drugs.

### Stopping rules

Stopping rules can be defined at the arm level. For example:

- The MTD can be considered reached if at least 9 patients have been evaluated in the trial arm, a dose level is the model’s recommendation for the next dose cohort two times in a row, and for this dose the posterior probability of targeted toxicity was at least 60% or at least 6 patients have been treated at this dose in the respective arm of the trial.
- A maximum sample size is reached in the trial arm, and then the MTD might *not* be declared.

### Reporting

Hypothetical data scenarios can be used to illustrate the model and the design. While the full-fledged `examine()` method might be difficult to implement as well as communicate, we can consider adding a new method e.g.`scenario()` that is a short cut for users to quickly get the model summary and next cohorts recommendation for a given hypothetical data scenario. This becomes more relevant now in the "hierarchical design" setting where multiple arms are included, and the data scenario can be defined for each arm separately.

Reporting of simulation results can start relatively simple and be located at the trial arm level, including e.g.:

- Mean number of patients treated at each dose combination
- Percentage of trials declaring an MTD per dose level
- Percentage of trial outcomes (underdose, target dose, overdose, stopped without reaching MTD)
- Mean, minimal, and maximal number of patients and patients with DLTs in simulated trials

### Probabilistic programming

The model can be implemented in a probabilistic programming language such as JAGS or Stan. The `decider` package by @Schroeter2023 already has an implementation in Stan, which can be used as a reference for the implementation in `crmPack`.
We will for the time being work with JAGS, as this is used elsewhere in the package. 

In case that we hit convergence issues with JAGS, we could consider adding Stan in particular for the "hierarchical design" setting, which has more parameters than the monotherapy case we have implemented so far.

One challenge here will be how to first specify, and second derive the JAGS model code from the user input across multiple arms in the "hierarchical design" setting. The user should be able to specify which arms are included in the design, and how they are connected in terms of shared parameters, and then the JAGS code should be derived from this specification. This will require some careful design of the user interface and the internal code structure.

## Existing material

### `crmPack` code

We have early code already in the [`dose_combinations`](https://github.com/openpharma/crmPack/tree/dose_combinations) branch (last changes happened 10 years ago!)

- The `Model` code is [here](https://github.com/openpharma/crmPack/blob/dose_combinations/R/Model-class.R#L2478). This uses the same 5-parameter BLRM model. It already foresees that the `LogisticLogNormal` model for the single agents could be replaced by other models, but does not yet support that. The code is implemented in JAGS. 
  - The prior specification is simpler than described above. There is no historical or concurrent monotherapy data included, and the prior means and covariances for the two mono models are fixed and not hierarchical. The interaction parameter $\eta$ has a simple (log) normal prior.
- The `Data` class is [here](https://github.com/openpharma/crmPack/blob/dose_combinations/R/Data-class.R#L331). It already is general for more than 2 drug combinations.
- The `NextBest` class is proposed [here](https://github.com/openpharma/crmPack/blob/dose_combinations/R/Rules-class.R#L156):
  - This rule will return the next best dose combinations according to the following algorithm: First, conditional on the first drug doses, the doses of the second drug which are admissible and maximize the target toxicity probabilities are recorded. Second, conditional on the second drug doses the same is done for the first drug doses. The union of both sets of drug combinations is returned.
- There is no `Design` class (or `Simulations` class) yet.
- There is prototype `Rcpp` code for speeding up the probability calculations based on samples [here](https://github.com/openpharma/crmPack/blob/main/design/devel/src/test.cpp).

### `decider` package

The `decider` package by @Schroeter2023 was specifically created for decision making in multiple-arm oncology dose escalation trials, including monotherapy and combination therapy arms. Historical trial data can be included as well. 

The model is described [here](https://boehringer-ingelheim.github.io/decider/reference/scenario_jointBLRM.html#details). It is based on the same 5-parameter BLRM approach as we describe here. It is implemented in Stan [here](https://github.com/Boehringer-Ingelheim/decider/blob/main/inst/stan/jointBLRM.stan).

The idea is to implement the most important functionality from the `decider` package in `crmPack`, in a structured way. (The `decider` package has grown organically over time from multiple studies it was applied to, and is not considered maintainable in the long term.)

## Implementation plan

The implementation in branch `936-combo-dev` has now two layers:

1. a fixed-prior two-drug combination model and data structure;
2. a hierarchical design framework that links named single-agent and combination arms through exchangeable parameter references: The hierarchical structure is encoded as named lists passed through `exchangeable_parameters`, and the model dynamically compiles the corresponding JAGS code.

### Fixed-prior two-drug combo layer

The fixed-prior combination layer is closest to the original 5-parameter BLRM idea.

- `DataCombo` stores two-drug dose-combination observations. Its `x` slot is an `n x 2` matrix, `doseGrid` is a named list of two dose grids, and `drugNames` records the two drug names. It also carries `backfilled` and optional efficacy `response` fields.
- `LogisticLogNormalCombo` combines two single-agent `LogisticLogNormal` models with the linear interaction term described above. The interaction parameter `eta` has an arm-specific normal or log-normal prior.
- `mcmc(DataCombo, LogisticLogNormalCombo, ...)` runs MCMC on the combination data and model, and returns `SamplesCombo` with the same structure as `Samples`, but with arm-specific parameter names.

This layer can be used either for a combination-only design with fixed priors, or be reused by the hierarchical model as the combination-arm likelihood.

### Hierarchical design layer

The more general hierarchical prototype keeps the original principle that each arm is still a standard `Design`, but wraps arms and compiles a joint model across them.

`DesignArm` is now a small concrete wrapper with only:

- `name`: unique arm name;
- `active`: whether the arm enrolls new patients. Inactive arms can represent historical arms;
- `design`: an ordinary `Design` object.

`HierarchicalData` stores a named list of arm data objects in `@arms`. Each arm may currently be a `Data` object or a `DataCombo` object. The implemented `update()` method targets one arm by name and leaves the other arms unchanged.

`HierarchicalModel` currently supports arm-specific models of class `LogisticLogNormal` and `LogisticLogNormalCombo`. It stores:

- `models_to_arms`: named list of arm-specific models;
- `parameter_pools`: named list of exchangeable parameter mappings.

The user-facing parameter references are deliberately simple. For a `LogisticLogNormal` arm, supported references are `"alpha0"` and `"alpha1"`. For a `LogisticLogNormalCombo` arm, supported references are `"alpha0[1]"`, `"alpha1[1]"`, `"alpha0[2]"`, and `"alpha1[2]"`. The interaction parameter `eta` remains arm-specific in the current prototype.

For example, a mono arm can share the first drug's intercept and slope with a combo arm as follows:

```{r eval=FALSE}
hmodel <- HierarchicalModel(
    mono = LogisticLogNormal(
        mean = c(-0.85, 1),
        cov = matrix(c(1, -0.5, -0.5, 1), nrow = 2),
        ref_dose = 10
    ),
    combo = LogisticLogNormalCombo(
        single_models = list(
            drug1 = LogisticLogNormal(
                mean = c(-0.85, 1),
                cov = matrix(c(1, -0.5, -0.5, 1), nrow = 2),
                ref_dose = 10
            ),
            drug2 = LogisticLogNormal(
                mean = c(-0.7, 0.8),
                cov = matrix(c(1.1, -0.3, -0.3, 0.9), nrow = 2),
                ref_dose = 20
            )
        ),
        gamma = 0,
        tau = 1
    ),
    exchangeable_parameters = list(
        drug1_intercept = list(
            mono = "alpha0",
            combo = "alpha0[1]"
        ),
        drug1_slope = list(
            mono = "alpha1",
            combo = "alpha1[1]"
        )
    )
)
```

Internally, `HierarchicalModel()` dynamically compiles:

- one joint JAGS likelihood by concatenating arm-specific likelihoods with arm-prefixed variable names;
- priors that keep unpooled parameters on their arm-specific fixed priors;
- exchangeable normal priors for pooled latent parameters;
- simple default hyperpriors for pool means and standard deviations.

The default hyperprior form mirrors the current prototype: intercept-like pool means use a normal prior centered at `logit(0.25)` with scale `2.5`, slope-like pool means use a normal prior centered at `0` with scale `0.7`, and pool standard deviations use log-normal priors with `kappa_hier = log(2) / 1.96`.

`HierarchicalDesign` is a constructor layer over named `DesignArm` objects. It derives:

- `@arms` from the supplied `DesignArm` objects;
- `@data` by collecting each `arm@design@data` into `HierarchicalData`;
- `@model` by collecting each `arm@design@model` into `HierarchicalModel`.

The corresponding API shape is:

```{r eval=FALSE}
hier_design <- HierarchicalDesign(
    DesignArm(
        name = "mono",
        active = TRUE,
        design = design_mono
    ),
    DesignArm(
        name = "combo",
        active = TRUE,
        design = design_combo
    ),
    exchangeable_parameters = list(
        drug1_intercept = list(
            mono = "alpha0",
            combo = "alpha0[1]"
        ),
        drug1_slope = list(
            mono = "alpha1",
            combo = "alpha1[1]"
        )
    )
)
```

Then MCMC is run on the derived joint data and model:

```{r eval=FALSE}
samples <- mcmc(
    data = hier_design@data,
    model = hier_design@model,
    options = mcmcOptions
)
```

The hierarchical MCMC method returns `HierarchicalSamples`, including a mapping from arm-level sample names such as `alpha0` and `alpha1` to the generated arm-prefixed sample names. This mapping is needed because generated JAGS variables include the arm names, e.g. `alpha0_mono`, `alpha1_mono`, `alpha0_combo`, `alpha1_combo`, and `eta_combo`.

The `simulate()` method for `HierarchicalDesign` runs trial simulations by running MCMC on the overall model at each iteration (i.e. cohort update across active arms), and then applies the arm-specific `NextBest`, `Increment` and `Stopping` rules to the corresponding arm-level samples and designs to obtain dose recommendations. The returned `HierarchicalSimulations` object can be summarized and reported at the arm level with the usual `summary()` method.
Also backfilling cohorts are already supported.

### Current limitations and open design questions

The current hierarchical prototype is much more concrete than the original sketch, but it is still experimental.

- Currently all arms open at the same time. We still need to allow separate start of arms. Here we need to think about a sensible API design. One direction could be to use something similar to the `Opening` class for the `Backfill` objects.
- `exchangeable_parameters` encodes symmetric exchangeability, not explicit source/borrower direction. The earlier idea that combo could borrow from mono without mono borrowing from combo is not yet represented as a first-class API concept. Note that this would be more complex and potentially require either data duplication or separate MCMC runs.
- Only `LogisticLogNormal` and `LogisticLogNormalCombo` arms are supported in `HierarchicalModel`.
- Only selected intercept and slope parameters can be pooled. The combo interaction parameter `eta` remains unpooled.
- A future `scenario()` helper remains attractive for quickly showing posterior summaries and next-dose recommendations across multiple arms.
