---
title: "Design: Table reporting and export for simulation studies"
subtitle: "True toxicity scenarios, hypothetical data scenarios, priors, and operating characteristics"
format:
  html:
    embed-resources: true
    toc: true
---

# Background

This design addresses [issue #966](https://github.com/openpharma/crmPack/issues/966):
we need standard, protocol-ready tables (and accompanying plots) that describe
simulation studies for dose escalation designs, suitable for inclusion in
clinical trial protocols and their appendices.

The requirements were collected in a team brainstorming (Bayer, Boehringer
Ingelheim, Merck examples; see the "Table discussion" notes) and fall into
four topics:

1. **True toxicity-scenario characterization**: assumed true DLT probability
   for each dose (or dose combination), as a table and as a plot.

1. **Hypothetical data-scenario evaluation**: observed patients and DLTs,
   current dose and its overdose probability, the next action recommended by
   the design, and posterior interval probabilities at the proposed next dose.
   These are evaluations of a design against user-supplied trial data, not the
   true toxicity scenarios used to generate simulation outcomes.

1. **Prior characterization**: summary of the prior distribution of the DLT
   probability at each dose

   - as a table (mean, SD, quantiles, interval probabilities), 
   - as a plot (central estimate with credible band, optional target interval overlay), and 
   - as a "spaghetti" plot of random prior draws.

1. **Operating characteristics**: 

   - an *overview* table with scenarios as
   columns and summary statistics as rows (PMDA-style, cf.
   [PMDA guidance, section 3.1, "Evaluation metrics for operating characteristics"](https://www.pmda.go.jp/files/000272426.pdf)), and
   - a *detail* table with per-dose-level results and scenarios as row groups,
   and simple per-dose barplots (selection probability, patients treated),
   optionally faceted by scenario.
   - Sponsor-specific needs on top of the PMDA core statistics: stop reason
   percentages, per-dose selection percentages, correct-dose-selected
   percentage, SD/quantiles/min–max of patient numbers, patients above target,
   trial duration, per-dose average patients and DLTs with optional SD, interval
   probabilities, and arm-specific results (sub-columns) for hierarchical
   designs.

Cross-cutting requirements:

- **User-level flexibility**: users must be able to 

  - add/remove statistics (rows) in the operating characteristics tables, 
  - choose display variants (e.g. "mean" vs. "mean (sd)", min–max, quantiles), and 
  - control row labels and column headers programmatically (to produce many table variants with
    different assumptions).

- **Export**: tables must render to markdown (the "easy" path) and be easy to
  copy/paste or export into Word documents.

- **Pagination**: a mechanism for automatic table breaks when there are too
  many columns (many scenarios or many dose levels).

# Current state

What already exists in `crmPack` and is relevant here:

- **Simulation results**: `Simulations` (and relatives) store per-simulation
  `data`, final `doses`, `fit`, `stop_report` (logical matrix with
  `report_label` column names), `stop_reasons`, and `additional_stats` (from
  the `derive` argument of `simulate()`).
- **Summaries**: `summary()` produces `GeneralSimulationsSummary` /
  `SimulationsSummary` objects storing e.g. `n_obs`, `n_above_target`,
  `dose_selected`, `prop_at_target`, `dose_most_selected`, `stop_report`,
  `mean_fit` (which includes the `truth` evaluated on the dose grid).
- **Printing**: `show()` methods on the summary classes `cat()` a fixed set of
  lines via an internal `Report` ref class ([Simulations-methods.R](../../R/Simulations-methods.R)),
  and return a one-row `data.frame` invisibly. This is neither
  machine-readable in a useful layout nor customizable.
- **Transient computations**: several statistics we need for the tables are
  currently computed *only* inside `show`/`plot` helpers and thrown away:
  per-dose selection percentages (inside `h_barplot_percentages()`), mean
  patients per dose (inside `h_plot_doses_tried()`), stop rule percentages
  (`h_calc_report_label_percentage()` in [helpers_simulations.R](../../R/helpers_simulations.R)).
- **Prior characterization**: the established idiom is
  `mcmc(empty_data, model, options)` followed by `fit()` (returning a
  `data.frame` with `dose`, `middle`, `lower`, `upper`) and
  `plot(samples, model, data)`. There is no tabular prior summary (mean, SD,
  arbitrary quantiles, interval probabilities) and no prior random-draws plot.
- **True toxicity-scenario characterization**: `truth` functions are passed to
  `simulate()` and `summary()`; `probFunction()` builds them from models. The
  true DLT curve is evaluated inside `summary()` but only retained in
  `SimulationsSummary@mean_fit$truth`. There is no scenario table/plot across
  multiple true toxicity scenarios.
- **Hypothetical data-scenario evaluation**: `scenario()` methods for `Design`,
  `DesignCombo`, `DADesign`, and `HierarchicalDesign` evaluate one
  user-supplied data set and currently return a plain named list containing the
  data, posterior samples and fit, next dose information, and stopping decision.
  The returned `next_best$probs` contains rule-specific posterior probabilities.
  The list has no class-level validity checks or methods, and there is no
  function that assembles multiple such results or renders them as a table.
- **Reporting layers**: two established conventions we should build on:
  - `tidy()` methods (broom-style, [helpers_broom.R](../../R/helpers_broom.R),
    `design/concise-tidy-design.Rmd`) returning tibbles classed
    `tbl_<ClassName>`.
  - `knit_print()` S3 methods ([helpers_knitr_*.R](../../R/helpers_knitr.R),
    `design/767-reporting-crmpack-objects/`) rendering objects in Quarto/R
    Markdown documents via `knitr::kable` + `kableExtra`.
  - Notably, there is not yet a `knit_print` or `tidy`-based table for any
    `*SimulationsSummary` class yet — that is part of the gap that this design fills.
- **Dependencies**: `knitr`, `kableExtra`, `dplyr`, `tibble` are in Imports;
  `broom`, `tidyr` in Suggests. No `gt`, `flextable`, or `rtables` yet.

# Design overview

We propose a strict **three-layer architecture**. Each layer is usable on its
own, and only the last layer knows anything about table rendering engines:

1. Statistics layer      
   
   - tidy tibbles from crmPack objects
   - exported accessor functions + tidy() methods
        
1. Assembly layer

   - report specification -> long "table skeleton" tibble
   - which statistics, which scenarios, labels, formats
        
1. Rendering layer

   - gt (rich HTML/Word) or knitr::kable (markdown)
   - knit_print() defaults, pagination, export helpers


The guiding principles:

- **Tibbles first.** Every table has a plain, documented tibble representation
  that users can post-process with dplyr and feed into *any* engine
  themselves. This is the ultimate flexibility escape hatch, and it makes the
  statistics unit-testable independently of formatting.
- **Statistics are pluggable.** The operating characteristics tables are
  driven by a registry of named statistics; users select built-ins by name,
  configure display variants, or supply their own functions with the same
  signature.
- **Rendering is thin and swappable.** We use `gt` as the rich rendering
  engine (see below), with `knitr::kable` markdown as the zero-extra-dependency
  fallback. Since layer 3 only consumes tibbles plus a small amount of layout
  metadata (spanners, row groups, footnotes), adding `flextable` later would
  be cheap if a strong Word-native need emerges.
- **Consistent naming.** All new exported reporting functions share a
  `report_` prefix (e.g. `report_oc_overview()`, `report_prob_summary()`), so
  they are easy to discover via autocomplete and clearly separated from the
  modeling/simulation API. New plot functions use a `plot_` prefix.

## Rendering engine choice

**Decision**: `gt` in **Suggests** as the primary rich renderer, with
`knitr::kable`-based markdown as the always-available fallback (uses only
existing Imports). All `format = "gt"` code paths check
`requireNamespace("gt")` and fail with an informative message. This keeps the
hard dependency footprint unchanged while covering both the "easy markdown"
and the "Word document" requirements.

Advantages: Modern, actively developed; spanners, row groups, footnotes, `fmt_*` formatting; exports HTML, LaTeX, RTF and **`.docx`** (`gt::gtsave("table.docx")`); `as_raw_html()` pastes cleanly into Word; widely known API

# Topic 1: True toxicity-scenario characterization

Here, "scenario" means an assumed true dose--toxicity curve used to generate
simulation outcomes. It does not mean the hypothetical observed data evaluated
by `scenario()` in Topic 2.

## New class: `ToxicityScenarios`

A lightweight container so that the set of assumed truths becomes a first-class
object that can be tabulated, plotted, and used to drive the per-scenario
`simulate()` calls:

```{r}
#| eval: false
ToxicityScenarios(
  scenarios = list(
    # named list of truth functions,
    "T1 (First)" = function(dose, ...) ..., # e.g. from probFunction()
    "T5 (Safe)" = probFunction(model, alpha0 = -3, alpha1 = 1)
  ),
  dose_grid = emptydata@doseGrid,
  target = c(0.2, 0.35) # optional target toxicity interval
)
```

Slots: `scenarios` (named list of functions), `dose_grid` (numeric),
`target` (numeric, optional). Contains `CrmPackClass`, with the usual
validity checks (functions accept a dose vector and return probabilities).

Methods:

- `tidy()` → tibble with columns `scenario`, `dose`, `prob`
  (long format; class `tbl_ToxicityScenarios`).
- `plot()` → the scenario definition plot: dose grid on x, DLT probability on
  y, one line/color per scenario (matches "Figure 10-1" in the notes);
  optional shaded target band via the `target` slot.
- `knit_print()` / `report_table()` → the scenario definition table: doses as
  columns, scenarios as rows (matches "Table 10-3"), with digit control.

For dose *combinations* (`ComboSimulations` world), the same class can later be
extended (`dose_grid` as a matrix / `DataCombo` grid, doses labelled
"d1 + d2"); the tidy format simply gains the second dose column. This is out of
scope for the first iteration but the tidy column naming should anticipate it.

## Storing the truth on the simulation objects

Several operating characteristics ("correct MTD", "overestimated MTD",
patients above target, the true P(DLT) row in the detail table) need the
`truth` function that generated the simulated trials. Instead of asking users
to pass the same function twice (once to `simulate()`, once to the reporting
functions) — with the risk of accidentally passing a different one —
`simulate()` will store the `truth` function it received in a new `truth` slot
on `GeneralSimulations` (and thereby on all subclasses). All downstream
reporting uses this stored truth.

Backward compatibility: the `truth` argument of `summary()` becomes optional,
defaulting to the stored slot; simulation objects saved with older `crmPack`
versions lack the slot, in which case the reporting functions ask for an
explicit `truth` via an informative error.

# Topic 2: Hypothetical data-scenario evaluation

This table reports what a design would do for several user-supplied snapshots
of trial data. It is deliberately based on the result of `scenario()` rather
than rerunning the model inside the reporting layer: the expensive stochastic
calculation happens once, and every exported table is traceable to the exact
fit and decision that were evaluated.

## New class: `DataScenario`

`scenario()` returns a `DataScenario` object representing one evaluated
hypothetical data scenario. This gives the result a stable contract and enables
class-specific `show()`, `tidy()`, and `knit_print()` methods. The distinction
in terminology is:

- a `Data`, `DataCombo`, `DataDA`, or `HierarchicalData` object is the
  user-supplied hypothetical data,
- a `DataScenario` is the complete evaluated result returned by `scenario()`,
  and
- a `ToxicityScenarios` object contains true dose--toxicity curves used for
  simulation.

`DataScenario` is a regular S4 class containing `CrmPackClass`, and
`scenario()` changes its `valueClass` from `"list"` to `"DataScenario"`.
Callers do not construct these objects directly. Its slots are `data`,
`samples`, `fit`, `dose_limit`, `next_best`, `next_dose`, `cohort_size`,
`placebo_cohort_size`, `stop`, `stop_report`, and `stop_reason`.

`data` and `samples` retain their existing crmPack objects. All remaining slots
are named lists with one element per arm. A non-hierarchical design uses one
internal arm named `"overall"`; renderers omit that artificial arm label.
Normalizing the representation here means downstream methods do not need one
code path for scalar single-design results and another for list-valued
hierarchical results. The validity function enforces identical arm names and
ordering across these slots, while allowing rule-dependent contents such as a
numeric or matrix `dose_limit` and a missing `next_best` for an unopened arm.

There is no compatibility path for the unclassed result lists returned by the
development version of `scenario()`. That API has not shipped on CRAN, so
`DataScenario` becomes its initial supported return contract.

The class validity function checks the compatibility of `data` and `samples`
and the shared per-arm structure of the decision slots. For hierarchical
results it also checks those names and their order against `HierarchicalData`.
It deliberately does not require a fixed column schema for the `probs` element
of each arm's `next_best`, because that schema belongs to the selected
`NextBest` rule.

Methods:

- `show()` gives a compact decision summary and does not print posterior
  samples.
- `tidy()` produces the normalized one-result tibble
  (`tbl_DataScenario`) described below, including one row for empty arms.
- `knit_print()` delegates to `report_data_scenarios()` for a one-scenario
  table.

`report_data_scenarios()` accepts either one `DataScenario` or a named list of
them. Other inputs are rejected with a clear type error. The reporter calls
the `tidy()` method for each result and binds those rows; the result-flattening
logic therefore has one implementation.

## Entry point and return value

For a multi-scenario table, the user supplies a named list of `DataScenario`
objects produced by `scenario()`:

```{r}
#| eval: false
hypothetical_results <- list(
  "1 (3 patients, 0 DLTs)" = scenario(
    design,
    data = data_scenario_1,
    mcmcOptions = options
  ),
  "2 (3 patients, 1 DLT)" = scenario(
    design,
    data = data_scenario_2,
    mcmcOptions = options
  )
)

report_data_scenarios(
  hypothetical_results,
  # For a NextBest rule that returns all three categories:
  probabilities = c("underdosing", "target", "overdose"),
  probability_labels = c(
    underdosing = "P(UDS)",
    target = "P(TDS)",
    overdose = "P(ODS)"
  ),
  format = c("tibble", "gt", "kable") # default "tibble"
)
```

`format = "tibble"` binds the individual `tbl_DataScenario` objects and returns
a tibble with class `tbl_DataScenarios`, with one row per scenario, arm, and
distinct observed dose (or dose combination). Its core columns are:

- `scenario` and `arm` (`arm` is omitted from the rendered table for a
  non-hierarchical design),
- `observed_dose`, `n_patients`, and `n_dlts`,
- `current_dose` and `current_overdose_prob`,
- `proposed_next_dose`, `cohort_size`, `stop`, and `stop_reason`, and
- one column for each requested posterior probability at the proposed next
  dose.

Dose columns retain a named numeric vector in a list-column and have a paired
character display column. Thus a single-agent dose remains numeric, a
`DataCombo` dose retains both drug components, and a `HierarchicalDesign` can
mix single-agent and combination arms without losing information. The default
display joins combination components with `" + "`; drug names, units, digits,
column labels, and the separator are configurable.

Observed counts are aggregated from `result@data` at each distinct dose. The
flattening helper starts from the existing per-arm `tidy()` methods instead of
duplicating extraction from the S4 slots. The current dose is the non-placebo
treatment dose in the latest cohort (the combination in the final row for
`DataCombo`); ambiguous latest-cohort data cause an error rather than an
arbitrary choice. Empty arms still get a row so that the scenario/arm layout
remains stable; unavailable values render as an em dash. For hierarchical
results, historical and not-yet-open arms are retained and their state is taken
from `result@stop` and `result@stop_reason` rather than inferred from missing
dose values.

## Decision probabilities and stopping

For each arm, the reporting function extracts probabilities from
`result@next_best[[arm]]$probs` at the current and proposed next doses. Known
crmPack spellings are normalized (for example, `target` and `target_prob`;
`overdose` and `overdose_prob`). It reuses `match_within_tolerance()` for
numeric dose matching. It does **not** calculate `1 - target - overdose` and label that value as
underdosing: `NextBest` intervals need not be contiguous, and some rules have
additional `excessive` or `unacceptable` categories. By default the table shows
the probability columns present in all evaluated arms; structurally unevaluated
historical or not-yet-open arms receive missing values and do not reduce that
set. If the user requests a missing column explicitly for an evaluated arm,
the function errors and lists the available columns instead of silently
changing the meaning of the table. A caller can include rule-specific
probability columns by name and relabel them for the protocol.

`scenario()` calculates `next_best` before evaluating `stopTrial()`. To avoid
presenting a dose as actionable after a stopping rule has fired, the rendered
table labels this column **Next action**: it shows the recommended dose and
cohort size when `stop` is false, and "Stop" with the stopping reason when
`stop` is true. The raw tibble keeps `proposed_next_dose`, `stop`, and
`stop_reason` separately. Posterior probabilities at a proposed next dose are
suppressed in the rendered table after a stop, but remain in the tibble for
auditability.

The `gt` renderer uses the screenshot's structure: row groups for scenarios,
arm rows within scenarios, an **Observed data** spanner over dose/patient/DLT
counts, and an **Interval probabilities of proposed next dose** spanner. When
an arm contains observations at more than one dose, the decision cells are
row-spanned in `gt` and displayed only on the first arm row in `kable`. The
classed tibble has a `knit_print()` method delegating to `report_table()`, so
the default tibble result can be placed directly in a Quarto chunk. Calling
`report_data_scenarios(..., format = "gt")` or `"kable"` renders immediately.

Tests cover single-agent, `DataCombo`, and `HierarchicalDesign` results;
multiple observed doses; empty, historical, and not-yet-open arms; stopped
designs; tolerant matching of numeric doses to probability grids; and clear
failure for requested probability categories that a `NextBest` result does
not provide. The reporter must not call `mcmc()`, `fit()`, or `nextBest()`.

# Topic 3: Prior characterization

All three deliverables operate on the established
`samples <- mcmc(empty_data, model, options)` idiom, so they automatically
work for **posterior** characterization too (non-empty data) — a nice bonus
for free.

## Prior definition table: `summary` of probability samples

New exported function:

```{r}
#| eval: false
report_prob_summary(
  samples,
  model,
  data,
  quantiles = c(0.025, 0.5, 0.975), # arbitrary quantiles
  mean = TRUE,
  sd = TRUE,
  intervals = c(0, 0.16, 0.33, 1) # optional: P(prob in [a, b)) columns
)
```

Returns a tibble (class `tbl_ProbSummary`) with one row per dose in
`data@doseGrid` and columns `dose`, `mean`, `sd`, `q2.5`, `q50`, `q97.5`,
`p_[0,0.16)`, `p_[0.16,0.33)`, `p_[0.33,1]`. This covers the Boehringer
"Summary of Prior" table exactly (interval probabilities + mean/SD/quantiles),
and degenerates to the Bayer variant by switching options off.

Implementation: reuse the `prob()` machinery already underlying `fit()`
(`fit()` itself is not sufficient since it fixes one central statistic and one
interval). A `knit_print`/`report_table` method renders it with a "Quantiles"
spanner and a "Probability of true DLT rate in ..." spanner, as in the
Boehringer example.

## Prior definition plot

Extend the existing `plot(samples, model, data)` method (or add a thin wrapper
if signature churn is a concern) with:

- `statistic = c("mean", "median")` for the central line,
- `level = 0.95` for the credible band,
- `target = NULL` — a scalar (horizontal target rate line) or length-2 vector
  (shaded target interval), drawn on top.

This matches the Bayer "Prior definition plot" (mean with 95% CI band and
target line).

## Prior random draws plot

New function:

```{r}
#| eval: false
plot_prior_draws(samples, model, data, n = 100, seed = NULL)
```

Draws `n` parameter vectors from the samples, evaluates the dose–toxicity
curve for each on (a fine grid over) the dose grid, and plots the spaghetti
curves (Bayer "Prior random draws plot" / "Figure 10-3"). Returns a `ggplot`.
A separate function (rather than a `type = "draws"` option on
`plot(samples, model, data)`) keeps the existing plot signature untouched and
the new functionality easy to discover.

# Topic 4: Operating characteristics

This is the largest part and the one needing the most flexibility.

## Data layer: promote transient statistics to accessors

First, factor the statistics that today live inside `show()`/`plot()` into
small exported functions returning tibbles (one job each, all unit-testable):

| Accessor | Content | Today |
|---|---|---|
| `report_dose_selection(sims)` | per-dose selection % incl. "none" | transient in `h_barplot_percentages()` |
| `report_patients_per_dose(sims)` | mean/SD/min/max patients per dose | transient in `h_plot_doses_tried()` |
| `report_dlts_per_dose(sims)` | mean/SD DLTs per dose | new |
| `report_stop_reasons(sims)` | % per `report_label`, grouped into categories (e.g. "toxicity", "max patients"), see below | partially `h_calc_report_label_percentage()` |
| `report_sample_size(sims)` | total n: mean, SD, min–max, quantiles | partially in summary |
| `report_mtd_classification(sims, target)` | % correct / overestimated / underestimated MTD, % stopped without MTD | partially (`prop_at_target`) |

Notes:

- "Correctly chosen MTD" = `truth(dose_selected)` inside `target`;
  "overestimated" = above the upper target bound. The `truth` comes from the
  slot stored on `sims` (see above); only `target` is passed.
- "% terminated due to toxicity" vs. "% max patients reached without MTD"
  cannot be derived from `stop_reasons` free text reliably; the design relies
  on `report_label`s plus a mapping of labels to categories, e.g.
  `stop_categories = list("Toxicity" = c("Lowest dose unsafe", "All doses toxic"), "Max patients" = "Futility")`.
  So that the PMDA tables work out of the box, `stop_categories = NULL`
  falls back to a **PMDA-style default mapping**: the `stop_report` assembly
  (`h_unpack_stopit()`) will additionally record the class of the stopping
  rule behind each column, and each bundled `Stopping*` class gets a default
  category (toxicity-related, maximum-patients, precision/other). A
  user-supplied `stop_categories` always overrides the defaults. We should
  still document prominently that giving stopping rules `report_label`s is a
  prerequisite for readable stop-reason reporting.
- `additional_stats` (from `simulate(..., derive = ...)`) are surfaced
  automatically as candidate rows, keeping the existing `derive` mechanism as
  the extension point for per-simulation quantities we do not compute.

## The statistics registry

The core flexibility mechanism. A statistic is a small S4 (or plain list-based
S3) object:

```{r}
#| eval: false
OCStat(
  label = "Mean number of patients", # row label (programmatic)
  fun = function(sims, summary, scenario) {
    # returns numeric vector
    mean(summary@n_obs)
  },
  format = "%.1f" # sprintf or function
)
```

`crmPack` ships a documented registry of built-ins covering the union of the
PMDA, BI, Bayer, and Merck requirements, selectable by name. Display variants
are expressed as distinct registry entries or via options, e.g.:

- `"n_pat_mean"`, `"n_pat_mean_sd"` ("13.8 (5.2)"), `"n_pat_minmax"`
  ("[3–30]"), `"n_pat_quantiles"` (10%/90%),
- `"n_dlt_mean"`, `"n_dlt_mean_sd"`, ...
- `"prop_stop_tox"`, `"prop_stop_max_patients"`, `"prop_stop_by_label"`
  (one row per `report_label` / category),
- `"prop_mtd_correct"`, `"prop_mtd_over"`, `"prop_mtd_none"`,
- `"dose_selection"` (one row per dose level + "none" + "highest dose safe"),
- `"n_above_target_mean_sd"`, `"duration_mean_sd"` (DA designs / scaled by
  number of cohorts),
- `"n_pat_per_dose"`, `"n_dlt_per_dose"` (one row per dose; `sd = TRUE` gives
  "mean (sd)" — the explicit Bayer request).

Users mix built-in names and custom `OCStat` objects in one `stats = list(...)`
argument, and can relabel any row (`label` override), which fulfils the
"add/remove statistics" and "programmatic headers/labels" requirements.

## Overview table (PMDA-style, scenarios as columns)

```{r}
#| eval: false
oc_summary <- SimulationsSummaryList(
  "Sc. 1 (DL1 toxic)" = sims1, # named Simulations arguments; each carries
  "Sc. 2 (MTD = DL1)" = sims2, # its truth in the stored slot
  target = c(0.2, 0.35),
  stop_categories = NULL # NULL -> PMDA-style defaults (see above)
)

report_oc_overview(
  oc_summary,
  stats = c(
    "n_pat_mean",
    "n_dlt_mean",
    "prop_stop_tox",
    "prop_mtd_correct",
    "prop_stop_max_patients",
    "prop_mtd_over"
  ),
  format = c("tibble", "gt", "kable") # default "tibble"
)
```

- The intermediate `SimulationsSummaryList` container (named list of
  per-scenario summaries + shared metadata) avoids recomputing summaries for
  each table variant and gives `plot()` a natural hook for scenario-faceted
  plots. A dedicated constructor is used rather than overloading `summary()`
  on a plain list, which would be unusual for S4 dispatch and harder to
  document. The per-scenario truths are taken from the `truth` slots of the
  supplied `Simulations` objects.
- `format = "tibble"` returns the long skeleton: `scenario`, `stat`, `label`,
  `value`, `formatted` — trivially reshaped by users.
- `format = "gt"` produces the rendered table: statistics as rows, scenarios
  as columns, an overall spanner like "Simulation Results (Assumptions: x, y,
  z)" supplied via a `title`/`spanner` argument (programmatic header
  requirement).
- Arm-specific sub-columns (BI requirement, `HierarchicalSimulations`): the
  skeleton gains an `arm` column and the gt renderer nests
  scenario → arm spanners. Designed for now, implemented as a follow-up.

## Detail table (per-dose, scenarios as row groups)

```{r}
#| eval: false
report_oc_detail(
  oc_summary,
  stats = c("n_dlt_per_dose", "n_pat_per_dose", "prop_selected", "true_prob"),
  sd = FALSE, # TRUE -> "mean (sd)" cells
  sum_column = TRUE, # SUM only for patient-count rows
  stop_columns = TRUE, # STOP / MAX columns only for the probability row
  format = c("tibble", "gt", "kable")
)
```

Matches the Merck detail table: columns are the dose levels (+ SUM/STOP/MAX
where applicable), row groups are the scenarios, rows within each group are
the selected per-dose statistics including the true P(DLT) row (evaluated from
the stored `truth` slots).

## Plots

Mostly re-uses existing machinery, generalized to multiple scenarios:

- `plot(oc_summary, type = "doseSelected")` and `type = "nPatients"` — per-dose
  barplots as today, but faceted by scenario (`facet_wrap(~ scenario)`).
- Existing single-`Simulations` plots stay untouched.

# Rendering and export layer

One generic entry point shared by all tables above:

```{r}
#| eval: false
report_table(x, format = c("gt", "kable"), ...)
# x: any of the tbl_* tibbles produced by layer 1/2 (they carry their class
#    and layout metadata as attributes: spanners, row groups, footnotes, digits)
```

- `knit_print()` methods for the new classes call `report_table()` with
  defaults, so bare objects in a Quarto chunk "just work" — consistent with
  the existing `helpers_knitr_*` convention.
- **Word export**: `report_table(x, format = "gt") |> gt::gtsave("table.docx")`
  or copy/paste of the rendered HTML; documented in the vignette.
- **Markdown**: `format = "kable"` emits pipe-table markdown using only
  existing Imports.
- **Pagination**: a helper `report_paginate(x, max_cols = 6, keep = "label")`
  splits a wide table into a list of tables repeating the row-label column(s);
  `report_table()` accepts the list and emits the parts in sequence. This
  covers "automatic page/table breaks if there are too many columns" for both
  many-scenario overview tables and many-dose detail tables.

# Dependencies

- Add `gt` to **Suggests** (guarded by `requireNamespace()`); it stays out of
  Imports deliberately to keep installation light, since the kable/markdown
  path is always available.
- No new Imports: the tibble/kable paths use `dplyr`, `tibble`, `knitr`,
  `kableExtra`. Any pivoting in the assembly layer is done with base R /
  `dplyr` only; `tidyr` stays in Suggests and would only be promoted to
  Imports if that turns out to be impractical during implementation.

# Implementation plan

Sliced so that each step is an independently reviewable PR with tests
(testthat + vdiffr for plots, snapshot tests for kable/gt output):

1. **Store the truth**: add the `truth` slot to `GeneralSimulations`,
   populate it in all `simulate()` methods, make the `truth` argument of
   `summary()` optional (defaulting to the stored slot), and record the
   stopping rule classes in the `stop_report` assembly (for the default
   `stop_categories`).
2. **Data-layer accessors**: factor per-dose selection %, patients/DLTs per
   dose, stop-report percentages, sample-size summaries, and MTD
   classification out of `show`/`plot` into exported, documented, tested
   `report_*` functions returning tibbles. Refactor `show`/`plot` to use them
   (no behavior change).
3. **`ToxicityScenarios`** class + `tidy`/`plot`/table methods.
4. **`DataScenario`** return class for `scenario()`, plus
   `report_data_scenarios()` and `tbl_DataScenario`/`tbl_DataScenarios` methods,
   including single-agent, combination, and hierarchical result flattening.
5. **`report_prob_summary()`** + prior plot extensions (`target` overlay,
   `statistic`/`level` options) + `plot_prior_draws()`.
6. **Statistics registry + `SimulationsSummaryList`** (`OCStat`, built-in
   registry, PMDA-style default `stop_categories`).
7. **`report_oc_overview()`** table (tibble + kable + gt formats, programmatic
   spanner/title).
8. **`report_oc_detail()`** + scenario-faceted summary plots.
9. **Rendering/export polish**: `report_table()` generic, `knit_print`
   methods, `report_paginate()`, Word-export documentation.
10. **Vignette** "Reporting simulation results" reproducing the hypothetical
    data-scenario, Bayer, BI, Merck, and PMDA example tables end-to-end.

Extensions explicitly deferred (but anticipated in the tidy column layout):
combo dose grids in true `ToxicityScenarios` and operating-characteristics
tables, arm sub-columns for `HierarchicalSimulations`, pseudo/dual simulation
specifics (efficacy columns), `DASimulations` duration statistics beyond
mean/SD. Combination and hierarchical `scenario()` results are already in
scope for the hypothetical data-scenario table described in Topic 2.
