---
title: "Design: Table reporting and export for simulation studies"
subtitle: "Scenario characterization, prior characterization, and operating characteristics"
format:
  html:
    embed-resources: true
    toc: true
---

# Background

This design addresses [issue #966](https://github.com/openpharma/crmPack/issues/966):
we need standard, protocol-ready tables (and accompanying plots) that describe
simulation studies for dose escalation designs, suitable for inclusion in
clinical trial protocols and their appendices.

The requirements were collected in a team brainstorming (Bayer, Boehringer
Ingelheim, Merck examples; see the "Table discussion" notes) and fall into
three topics:

1. **Scenario characterization**: assumed true DLT probability for each dose
   (or dose combination), as a table and as a plot.

1. **Prior characterization**: summary of the prior distribution of the DLT
   probability at each dose

   - as a table (mean, SD, quantiles, interval probabilities), 
   - as a plot (central estimate with credible band, optional target interval overlay), and 
   - as a "spaghetti" plot of random prior draws.

1. **Operating characteristics**: 

   - an *overview* table with scenarios as
   columns and summary statistics as rows (PMDA-style, cf.
   [PMDA guidance, section 3.1, "Evaluation metrics for operating characteristics"](https://www.pmda.go.jp/files/000272426.pdf)), and
   - a *detail* table with per-dose-level results and scenarios as row groups,
   and simple per-dose barplots (selection probability, patients treated),
   optionally faceted by scenario.
   - Sponsor-specific needs on top of the PMDA core statistics: stop reason
   percentages, per-dose selection percentages, correct-dose-selected
   percentage, SD/quantiles/min–max of patient numbers, patients above target,
   trial duration, per-dose average patients and DLTs with optional SD, interval
   probabilities, and arm-specific results (sub-columns) for hierarchical
   designs.

Cross-cutting requirements:

- **User-level flexibility**: users must be able to 

  - add/remove statistics (rows) in the operating characteristics tables, 
  - choose display variants (e.g. "mean" vs. "mean (sd)", min–max, quantiles), and 
  - control row labels and column headers programmatically (to produce many table variants with
    different assumptions).

- **Export**: tables must render to markdown (the "easy" path) and be easy to
  copy/paste or export into Word documents.

- **Pagination**: a mechanism for automatic table breaks when there are too
  many columns (many scenarios or many dose levels).

# Current state

What already exists in `crmPack` and is relevant here:

- **Simulation results**: `Simulations` (and relatives) store per-simulation
  `data`, final `doses`, `fit`, `stop_report` (logical matrix with
  `report_label` column names), `stop_reasons`, and `additional_stats` (from
  the `derive` argument of `simulate()`).
- **Summaries**: `summary()` produces `GeneralSimulationsSummary` /
  `SimulationsSummary` objects storing e.g. `n_obs`, `n_above_target`,
  `dose_selected`, `prop_at_target`, `dose_most_selected`, `stop_report`,
  `mean_fit` (which includes the `truth` evaluated on the dose grid).
- **Printing**: `show()` methods on the summary classes `cat()` a fixed set of
  lines via an internal `Report` ref class ([Simulations-methods.R](../../R/Simulations-methods.R)),
  and return a one-row `data.frame` invisibly. This is neither
  machine-readable in a useful layout nor customizable.
- **Transient computations**: several statistics we need for the tables are
  currently computed *only* inside `show`/`plot` helpers and thrown away:
  per-dose selection percentages (inside `h_barplot_percentages()`), mean
  patients per dose (inside `h_plot_doses_tried()`), stop rule percentages
  (`h_calc_report_label_percentage()` in [helpers_simulations.R](../../R/helpers_simulations.R)).
- **Prior characterization**: the established idiom is
  `mcmc(empty_data, model, options)` followed by `fit()` (returning a
  `data.frame` with `dose`, `middle`, `lower`, `upper`) and
  `plot(samples, model, data)`. There is no tabular prior summary (mean, SD,
  arbitrary quantiles, interval probabilities) and no prior random-draws plot.
- **Scenario characterization**: `truth` functions are passed to `simulate()`
  and `summary()`; `probFunction()` builds them from models. The true DLT
  curve is evaluated inside `summary()` but only retained in
  `SimulationsSummary@mean_fit$truth`. There is no scenario table/plot across
  multiple scenarios.
- **Reporting layers**: two established conventions we should build on:
  - `tidy()` methods (broom-style, [helpers_broom.R](../../R/helpers_broom.R),
    `design/concise-tidy-design.Rmd`) returning tibbles classed
    `tbl_<ClassName>`.
  - `knit_print()` S3 methods ([helpers_knitr_*.R](../../R/helpers_knitr.R),
    `design/767-reporting-crmpack-objects/`) rendering objects in Quarto/R
    Markdown documents via `knitr::kable` + `kableExtra`.
  - Notably, there is not yet a `knit_print` or `tidy`-based table for any
    `*SimulationsSummary` class yet — that is part of the gap that this design fills.
- **Dependencies**: `knitr`, `kableExtra`, `dplyr`, `tibble` are in Imports;
  `broom`, `tidyr` in Suggests. No `gt`, `flextable`, or `rtables` yet.

# Design overview

We propose a strict **three-layer architecture**. Each layer is usable on its
own, and only the last layer knows anything about table rendering engines:

1. Statistics layer      
   
   - tidy tibbles from crmPack objects
   - exported accessor functions + tidy() methods
        
1. Assembly layer

   - report specification -> long "table skeleton" tibble
   - which statistics, which scenarios, labels, formats
        
1. Rendering layer

   - gt (rich HTML/Word) or knitr::kable (markdown)
   - knit_print() defaults, pagination, export helpers


The guiding principles:

- **Tibbles first.** Every table has a plain, documented tibble representation
  that users can post-process with dplyr and feed into *any* engine
  themselves. This is the ultimate flexibility escape hatch, and it makes the
  statistics unit-testable independently of formatting.
- **Statistics are pluggable.** The operating characteristics tables are
  driven by a registry of named statistics; users select built-ins by name,
  configure display variants, or supply their own functions with the same
  signature.
- **Rendering is thin and swappable.** We recommend `gt` as the rich rendering
  engine (see below), with `knitr::kable` markdown as the zero-extra-dependency
  fallback. Since layer 3 only consumes tibbles plus a small amount of layout
  metadata (spanners, row groups, footnotes), adding `flextable` later would
  be cheap if a strong Word-native need emerges.

## Rendering engine choice

**Recommendation**: `gt` in **Suggests** as the primary rich renderer, with
`knitr::kable`-based markdown as the always-available fallback (uses only
existing Imports). All `format = "gt"` code paths check
`requireNamespace("gt")` and fail with an informative message. This keeps the
hard dependency footprint unchanged while covering both the "easy markdown"
and the "Word document" requirements.

Advantages: Modern, actively developed; spanners, row groups, footnotes, `fmt_*` formatting; exports HTML, LaTeX, RTF and **`.docx`** (`gt::gtsave("table.docx")`); `as_raw_html()` pastes cleanly into Word; widely known API

# Topic 1: Scenario characterization

## New class: `ToxicityScenarios`

A lightweight container so that the set of assumed truths becomes a first-class
object that can be tabulated, plotted, and passed to the operating
characteristics reporting (which needs the truths again for "correct MTD"
classification):

```{r}
#| eval: false
ToxicityScenarios(
    scenarios = list( # named list of truth functions,
        "T1 (First)" = function(dose, ...) ..., # e.g. from probFunction()
        "T5 (Safe)"  = probFunction(model, alpha0 = -3, alpha1 = 1)
    ),
    dose_grid = emptydata@doseGrid,
    target = c(0.2, 0.35) # optional target toxicity interval
)
```

Slots: `scenarios` (named list of functions), `dose_grid` (numeric),
`target` (numeric, optional). Contains `CrmPackClass`, with the usual
validity checks (functions accept a dose vector and return probabilities).

Methods:

- `tidy()` → tibble with columns `scenario`, `dose`, `prob`
  (long format; class `tbl_ToxicityScenarios`).
- `plot()` → the scenario definition plot: dose grid on x, DLT probability on
  y, one line/color per scenario (matches "Figure 10-1" in the notes);
  optional shaded target band via the `target` slot.
- `knit_print()` / `render_table()` → the scenario definition table: doses as
  columns, scenarios as rows (matches "Table 10-3"), with digit control.

For dose *combinations* (`ComboSimulations` world), the same class can later be
extended (`dose_grid` as a matrix / `DataCombo` grid, doses labelled
"d1 + d2"); the tidy format simply gains the second dose column. This is out of
scope for the first iteration but the tidy column naming should anticipate it.

# Topic 2: Prior characterization

All three deliverables operate on the established
`samples <- mcmc(empty_data, model, options)` idiom, so they automatically
work for **posterior** characterization too (non-empty data) — a nice bonus
for free.

## Prior definition table: `summary` of probability samples

New exported function (working title):

```{r}
#| eval: false
prob_summary(
    samples, model, data,
    quantiles = c(0.025, 0.5, 0.975), # arbitrary quantiles
    mean = TRUE,
    sd = TRUE,
    intervals = c(0, 0.16, 0.33, 1) # optional: P(prob in [a, b)) columns
)
```

Returns a tibble (class `tbl_ProbSummary`) with one row per dose in
`data@doseGrid` and columns `dose`, `mean`, `sd`, `q2.5`, `q50`, `q97.5`,
`p_[0,0.16)`, `p_[0.16,0.33)`, `p_[0.33,1]`. This covers the Boehringer
"Summary of Prior" table exactly (interval probabilities + mean/SD/quantiles),
and degenerates to the Bayer variant by switching options off.

Implementation: reuse the `prob()` machinery already underlying `fit()`
(`fit()` itself is not sufficient since it fixes one central statistic and one
interval). A `knit_print`/`render_table` method renders it with a "Quantiles"
spanner and a "Probability of true DLT rate in ..." spanner, as in the
Boehringer example.

## Prior definition plot

Extend the existing `plot(samples, model, data)` method (or add a thin wrapper
if signature churn is a concern) with:

- `statistic = c("mean", "median")` for the central line,
- `level = 0.95` for the credible band,
- `target = NULL` — a scalar (horizontal target rate line) or length-2 vector
  (shaded target interval), drawn on top.

This matches the Bayer "Prior definition plot" (mean with 95% CI band and
target line).

## Prior random draws plot

New function:

```{r}
#| eval: false
plot_prior_draws(samples, model, data, n = 100, seed = NULL)
```

Draws `n` parameter vectors from the samples, evaluates the dose–toxicity
curve for each on (a fine grid over) the dose grid, and plots the spaghetti
curves (Bayer "Prior random draws plot" / "Figure 10-3"). Returns a `ggplot`.
Alternatively this could be `plot(samples, model, data, type = "draws")`; the
decision should follow whichever signature the prior definition plot lands on
(open question below).

# Topic 3: Operating characteristics

This is the largest part and the one needing the most flexibility.

## Data layer: promote transient statistics to accessors

First, factor the statistics that today live inside `show()`/`plot()` into
small exported functions returning tibbles (one job each, all unit-testable):

| Accessor (working names) | Content | Today |
|---|---|---|
| `dose_selection_table(sims)` | per-dose selection % incl. "none" | transient in `h_barplot_percentages()` |
| `patients_per_dose_table(sims)` | mean/SD/min/max patients per dose | transient in `h_plot_doses_tried()` |
| `dlts_per_dose_table(sims)` | mean/SD DLTs per dose | new |
| `stop_report_table(sims)` | % per `report_label`, with user-defined grouping of labels into categories (e.g. "toxicity", "max patients") | partially `h_calc_report_label_percentage()` |
| `sample_size_table(sims)` | total n: mean, SD, min–max, quantiles | partially in summary |
| `mtd_classification_table(sims, truth, target)` | % correct / overestimated / underestimated MTD, % stopped without MTD | partially (`prop_at_target`) |

Notes:

- "Correctly chosen MTD" = `truth(dose_selected)` inside `target`;
  "overestimated" = above the upper target bound. This needs `truth` and
  `target` per scenario — which is why `ToxicityScenarios` carries both.
- "% terminated due to toxicity" vs. "% max patients reached without MTD"
  cannot be derived from `stop_reasons` free text reliably; the design relies
  on `report_label`s plus a user-supplied mapping, e.g.
  `stop_categories = list("Toxicity" = c("Lowest dose unsafe", "All doses toxic"), "Max patients" = "Futility")`.
  We should document prominently that giving stopping rules `report_label`s is
  a prerequisite for stop-reason reporting.
- `additional_stats` (from `simulate(..., derive = ...)`) are surfaced
  automatically as candidate rows, keeping the existing `derive` mechanism as
  the extension point for per-simulation quantities we do not compute.

## The statistics registry

The core flexibility mechanism. A statistic is a small S4 (or plain list-based
S3) object:

```{r}
#| eval: false
OCStat(
    label = "Mean number of patients", # row label (programmatic)
    fun = function(sims, summary, scenario) { # returns numeric vector
        mean(summary@n_obs)
    },
    format = "%.1f" # sprintf or function
)
```

`crmPack` ships a documented registry of built-ins covering the union of the
PMDA, BI, Bayer, and Merck requirements, selectable by name. Display variants
are expressed as distinct registry entries or via options, e.g.:

- `"n_pat_mean"`, `"n_pat_mean_sd"` ("13.8 (5.2)"), `"n_pat_minmax"`
  ("[3–30]"), `"n_pat_quantiles"` (10%/90%),
- `"n_dlt_mean"`, `"n_dlt_mean_sd"`, ...
- `"prop_stop_tox"`, `"prop_stop_max_patients"`, `"prop_stop_by_label"`
  (one row per `report_label` / category),
- `"prop_mtd_correct"`, `"prop_mtd_over"`, `"prop_mtd_none"`,
- `"dose_selection"` (one row per dose level + "none" + "highest dose safe"),
- `"n_above_target_mean_sd"`, `"duration_mean_sd"` (DA designs / scaled by
  number of cohorts),
- `"n_pat_per_dose"`, `"n_dlt_per_dose"` (one row per dose; `sd = TRUE` gives
  "mean (sd)" — the explicit Bayer request).

Users mix built-in names and custom `OCStat` objects in one `stats = list(...)`
argument, and can relabel any row (`label` override), which fulfils the
"add/remove statistics" and "programmatic headers/labels" requirements.

## Overview table (PMDA-style, scenarios as columns)

```{r}
#| eval: false
oc_summary <- summary(
    list(
        "Sc. 1 (DL1 toxic)" = sims1, # named list of Simulations
        "Sc. 2 (MTD = DL1)" = sims2
    ),
    scenarios = my_scenarios, # ToxicityScenarios: truths + target
    stop_categories = list(...)
) # -> "SimulationsSummaryList"

oc_table(
    oc_summary,
    stats = c(
        "n_pat_mean", "n_dlt_mean",
        "prop_stop_tox", "prop_mtd_correct",
        "prop_stop_max_patients", "prop_mtd_over"
    ),
    format = c("tibble", "gt", "kable") # default "tibble"
)
```

- The intermediate `SimulationsSummaryList` container (named list of
  per-scenario summaries + shared metadata) avoids recomputing summaries for
  each table variant and gives `plot()` a natural hook for scenario-faceted
  plots.
- `format = "tibble"` returns the long skeleton: `scenario`, `stat`, `label`,
  `value`, `formatted` — trivially reshaped by users.
- `format = "gt"` produces the rendered table: statistics as rows, scenarios
  as columns, an overall spanner like "Simulation Results (Assumptions: x, y,
  z)" supplied via a `title`/`spanner` argument (programmatic header
  requirement).
- Arm-specific sub-columns (BI requirement, `HierarchicalSimulations`): the
  skeleton gains an `arm` column and the gt renderer nests
  scenario → arm spanners. Designed for now, implemented as a follow-up.

## Detail table (per-dose, scenarios as row groups)

```{r}
#| eval: false
oc_detail_table(
    oc_summary,
    stats = c("n_dlt_per_dose", "n_pat_per_dose", "prop_selected", "true_prob"),
    sd = FALSE, # TRUE -> "mean (sd)" cells
    sum_column = TRUE, # SUM only for patient-count rows
    stop_columns = TRUE, # STOP / MAX columns only for the probability row
    format = c("tibble", "gt", "kable")
)
```

Matches the Merck detail table: columns are the dose levels (+ SUM/STOP/MAX
where applicable), row groups are the scenarios, rows within each group are
the selected per-dose statistics including the true P(DLT) row (taken from the
`ToxicityScenarios`).

## Plots

Mostly re-uses existing machinery, generalized to multiple scenarios:

- `plot(oc_summary, type = "doseSelected")` and `type = "nPatients"` — per-dose
  barplots as today, but faceted by scenario (`facet_wrap(~ scenario)`).
- Existing single-`Simulations` plots stay untouched.

# Rendering and export layer

One generic entry point shared by all tables above:

```{r}
#| eval: false
render_table(x, format = c("gt", "kable"), ...)
# x: any of the tbl_* tibbles produced by layer 1/2 (they carry their class
#    and layout metadata as attributes: spanners, row groups, footnotes, digits)
```

- `knit_print()` methods for the new classes call `render_table()` with
  defaults, so bare objects in a Quarto chunk "just work" — consistent with
  the existing `helpers_knitr_*` convention.
- **Word export**: `render_table(x, format = "gt") |> gt::gtsave("table.docx")`
  or copy/paste of the rendered HTML; documented in the vignette.
- **Markdown**: `format = "kable"` emits pipe-table markdown using only
  existing Imports.
- **Pagination**: a helper `paginate_columns(x, max_cols = 6, keep = "label")`
  splits a wide table into a list of tables repeating the row-label column(s);
  `render_table()` accepts the list and emits the parts in sequence. This
  covers "automatic page/table breaks if there are too many columns" for both
  many-scenario overview tables and many-dose detail tables.

# Dependencies

- Add `gt` to **Suggests** (guarded by `requireNamespace()`).
- No new Imports: the tibble/kable paths use `dplyr`, `tibble`, `tidyr`
  (Suggests → consider promoting to Imports since the assembly layer will
  need pivoting), `knitr`, `kableExtra`.

# Implementation plan

Sliced so that each step is an independently reviewable PR with tests
(testthat + vdiffr for plots, snapshot tests for kable/gt output):

1. **Data-layer accessors**: factor per-dose selection %, patients/DLTs per
   dose, stop-report percentages, sample-size summaries, and MTD
   classification out of `show`/`plot` into exported, documented, tested
   functions returning tibbles. Refactor `show`/`plot` to use them (no
   behavior change).
2. **`ToxicityScenarios`** class + `tidy`/`plot`/table methods.
3. **`prob_summary()`** + prior plot extensions (`target` overlay,
   `statistic`/`level` options) + `plot_prior_draws()`.
4. **Statistics registry + `SimulationsSummaryList`** (`OCStat`, built-in
   registry, multi-scenario `summary()`).
5. **`oc_table()`** overview table (tibble + kable + gt formats, programmatic
   spanner/title).
6. **`oc_detail_table()`** + scenario-faceted summary plots.
7. **Rendering/export polish**: `render_table()` generic, `knit_print`
   methods, `paginate_columns()`, Word-export documentation.
8. **Vignette** "Reporting simulation results" reproducing the Bayer, BI,
   Merck, and PMDA example tables end-to-end.

Extensions explicitly deferred (but anticipated in the tidy column layout):
combo dose grids in `ToxicityScenarios`/tables, arm sub-columns for
`HierarchicalSimulations`, pseudo/dual simulation specifics (efficacy
columns), `DASimulations` duration statistics beyond mean/SD.

# Open questions

1. **Naming**: `oc_table()` / `oc_detail_table()` / `prob_summary()` /
   `render_table()` are working titles — do we want a common prefix (e.g.
   `report_*` or `tabulate_*`), or S4 generics (`summary` on a list is
   somewhat unusual; a dedicated constructor like
   `SimulationsSummaryList(...)` may be cleaner than overloading `summary`)?

   - A `report_` prefix would be helpful

2. **Where should `truth` live long term?** Storing the evaluated truth (and
   the truth function) on the `Simulations` object at `simulate()` time would
   remove the need to pass scenarios twice (once to `simulate`, once to the
   reporting). That is a bigger change to a central class — worth it?

   - Yes good idea we can store it in simulations object, and then use that
   stored attribute/property in the downstream functions. I don't see the point
   of passing another different function as the `truth`.

3. **`tidyr` promotion** from Suggests to Imports for the assembly layer, or
   avoid pivoting with base/`dplyr` only?

   - If we really need `tidyr` we can promote to Imports, otherwise work with base/`dplyr` only.

4. **Prior draws plot signature**: new `type = "draws"` on
   `plot(samples, model, data)` vs. a separate `plot_prior_draws()` function.

   - A separate `plot_prior_draws()` function sounds better.

5. **gt in Suggests vs. Imports**: Suggests keeps installation light but means
   the "nice" output needs an extra install; is that acceptable for the
   primary rendering path?

   - Yes that is acceptable, please keep it in Suggests.

6. **Should `stop_categories` mappings get standard defaults** (e.g. by
   convention on `report_label`s), so PMDA tables work out of the box for the
   bundled designs?

   - Yes, PMDA style defaults would be helpful.
